---
title: "VPA Dose Experiment"
author: "Darya Akimova"
date: "November 2018"
output: 
  html_document:
    toc: true
    toc_float:
      smooth_scroll: false
    number_sections: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```


# Setup

## Packages

```{r packages, comment=NA, message = FALSE}
library(tidyverse)
library(cowplot)
library(GGally)
library(heatmaply)
library(sva)
library(limma)
library(biobroom)
library(ggridges)
```


## Data

### Metabolite Abundances

```{r abundance_import, comment=NA}
vpa.cell.neg.raw <- read_csv("./data/abundances/vpa_dose_exp_cells_negmode_abundances.csv")
vpa.cell.pos.raw <- read_csv("./data/abundances/vpa_dose_exp_cells_posmode_abundances.csv")
```

### Compound Information

```{r compound_info_import, comment=NA}
vpa.cell.neg.compound.info <- read_csv("./data/compound_info/vpa_dose_exp_cells_negmode_compound_info.csv")
vpa.cell.pos.compound.info <- read_csv("./data/compound_info/vpa_dose_exp_cells_posmode_compound_info.csv")
```

## Functions

```{r functions, comment=NA}
MissingPerSamplePlot <- function(raw.data, start.string) {
  # Counts the number of missing/NA values per sample and
  # percent compounds missing out of total number of compounds per sample
  # Then passes the result into a vertical bar plot, where each 
  # bar represents a single sample and the size of the bar 
  # is the % of compounds missing
  
  counted.na <- raw.data %>%
  select(starts_with(start.string)) %>% 
  mutate(
    count.na = apply(., 1, function(x) sum(is.na(x))),
    percent.na = (count.na / ncol(raw.data %>% select(starts_with(start.string)))) * 100
    ) %>%
  dplyr::select(count.na, percent.na) %>%
  bind_cols(
    raw.data %>% 
      select(Samples, Group)
      ) %>% 
  arrange(percent.na) %>% 
  mutate(f.order = factor(Samples, levels = Samples))
counted.na %>% 
  ggplot(aes(x = f.order, y = percent.na, fill = Group)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 20, color = "gray", size = 1, alpha = 0.8) +
  coord_flip()+
  xlab("Samples") +
  ylab("Percent missing values in sample") +
  theme(axis.text.y = element_text(size = 6)) 
}
MissingPerCompound <- function(raw.data, start.string){
  # Function to count in how many experimental samples each compound is missing
  # Also calculates the % missing:
  # (# NA per compound across all experimental samples) * 100 / (tot num of samples)
  
  raw.data %>% 
  filter(Group == "sample") %>% 
  select(Samples, starts_with(start.string)) %>% 
  gather(key = "Compound", value = "Abundance", -Samples) %>% 
  group_by(Compound) %>% 
  summarise(
    na_count = sum(is.na(Abundance)),
    n_samples = n(),
    percent_na = (na_count * 100 / n_samples)
    ) %>% 
  filter(na_count > 0) %>% 
  arrange(desc(percent_na))
}
ReplaceNAwMinLogTransformSingle <- function(raw.dataframe, start.prefix) {
  # Function to replace any NAs in each column with the minimum for that column, 
  # separately for each sample type.
  # NA in negative control samples are replaced by 2.
  # Then data is log2 transformed
  
  smpls <- raw.dataframe %>%
    filter(Group == "sample") %>% 
    dplyr::select(starts_with(start.prefix))
  smpls.min <- lapply(smpls, min, na.rm = TRUE)
  # replace the missing values in the real samples with the minimum of the samples
  # then take the log
  smpls.noNA <- raw.dataframe  %>%
    filter(Group == "sample") %>% 
    dplyr::select(Samples:Dose_uM) %>%
    bind_cols(
      smpls %>%
        replace_na(replace = smpls.min) %>% 
        log2()
      )
  # replace the missing values in solv and empty samples with 2 - for PCA analysis
  # then take the log
  other.min <- setNames(
    as.list(
      rep(2, ncol(
        raw.dataframe %>% 
          dplyr::select(starts_with(start.prefix))))
      ),
    colnames(raw.dataframe %>% dplyr::select(starts_with(start.prefix)))
                )
  other.num.log <- raw.dataframe %>%
    filter(Group != "sample" & Group != "mix") %>% 
    dplyr::select(Samples:Dose_uM) %>%
    bind_cols(
      raw.dataframe %>% 
        filter(Group != "sample" & Group != "mix") %>% 
        dplyr::select(starts_with(start.prefix)) %>%
        replace_na(replace = other.min) %>% 
        log2()
      )
  # combine them together back into one data frame
  all.noNA <- smpls.noNA %>% 
    bind_rows(other.num.log)
}
HeatmapPrepAlt <- function(raw.data, start.prefix){
  # function for preparing dara for heatmap viz
  x <- raw.data %>% 
    select(starts_with(start.prefix)) %>% 
    scale(center = TRUE, scale = TRUE) %>% 
    as.matrix() 
  row.names(x) <- raw.data$Samples
  return(x)
}
```


# Data Exploration

## Mass vs Retention Time

Q: What are the distributions of compound masses and retention times?

```{r mass_v_RT_plots, fig.height=5, fig.width=8}
full.vpa.cmpnd <- vpa.cell.neg.compound.info %>% 
  mutate(Set = "Cells / Neg") %>% 
  bind_rows(vpa.cell.pos.compound.info %>% mutate(Set = "Cells / Pos"))
full.vpa.cmpnd %>% 
  ggplot(aes(x = rt, y = mass)) +
  geom_point(size = 3, alpha = 0.5) +
  xlab("Retention Time (min)") +
  ylab("Mass (Da)") +
  ggtitle("Mass v RT\nVPA Dose Experiment") +
  ylim(0, 1000)
full.vpa.cmpnd %>% 
  ggplot(aes(x = rt, y = mass, color = Set)) +
  geom_point(size = 3, alpha = 0.8) +
  xlab("Retnetion Time (min)") +
  ylab("Mass (Da)") +
  ggtitle("Mass v RT\nVPA Dose Experiment") +
  facet_wrap(~ Set) +
  ylim(0, 1000)
```

Q: Which compounds were found in one or more of the data types?

```{r cmnpd_mode_match, fig.height= 5, fig.width=8, comment=NA}
vpa.cell.cmpnd.join <- vpa.cell.neg.compound.info %>% 
  inner_join(vpa.cell.pos.compound.info, by = "cas_id", suffix = c(".c.n", ".c.p")) %>% 
  select(
    contains("cas_id"), contains("short"), 
    contains("full"), starts_with("formula"), 
    starts_with("mass"), starts_with("rt")
    )
# compound names - found in pos and neg mode / cells 
print(vpa.cell.cmpnd.join$compound_full.c.n)
# percent of cell / neg compounds found in cell / pos 
round(nrow(vpa.cell.cmpnd.join) * 100 / nrow(vpa.cell.neg.compound.info), 1)
# percent of cell / neg compounds found in cell / pos 
round(nrow(vpa.cell.cmpnd.join) * 100 / nrow(vpa.cell.pos.compound.info), 1)
# any mass inconsistencies?
vpa.cell.cmpnd.join %>% 
  select(contains("mass")) %>% 
  ggpairs()
# any rt inconsistencies?
vpa.cell.cmpnd.join %>% 
  select(starts_with("rt")) %>% 
  ggpairs()
```


## Missing Values

```{r missing_per_sample_plots, comment=NA}
MissingPerSamplePlot(vpa.cell.neg.raw, "VPAnC") +
  ggtitle("Missing Per Sample\nVPA Dose Experiment / Cells / Neg Mode")
MissingPerSamplePlot(vpa.cell.pos.raw, "VPApC") +
  ggtitle("Missing Per Sample\nVPA Dose Experiment / Cells / Pos Mode")
```


Q: Are any of the compounds more than 20% missing in the experimental sample group? If there are any, they will be excluded from analysis. 

```{r percent_missing_per_compound, comment=NA}
vpa.cell.neg.raw %>% 
  MissingPerCompound("VPAnC") %>% 
  filter(percent_na > 20)
(vpa.cell.pos.cmpnd.excl <- vpa.cell.pos.raw %>% 
  MissingPerCompound("VPApC") %>% 
  filter(percent_na > 20))
```


## Negative Controls and Compound Elimination

### Cells / Negative Mode

```{r solv_cell_neg, comment=NA}
vpa.cell.neg.raw.grp.mean <- vpa.cell.neg.raw %>% 
  group_by(Group) %>% 
  summarize_at(vars(matches("VPAnC")), mean, na.rm = TRUE) %>% 
  gather(key = "Compound", value = "Group_mean_abun", -Group)
vpa.cell.neg.raw.grp.mean %>% 
  ggplot(aes(log2(Group_mean_abun), color = Group)) +
  geom_density(size = 2, alpha = 0.8) +
  ggtitle("Distribution of compound means\nVPA Dose Exp / Cells / Negative Mode\nGrouped by sample type")
vpa.cell.neg.raw.grp.mean.order <- vpa.cell.neg.raw.grp.mean %>% 
  filter(Group == "sample") %>% 
  arrange(Group_mean_abun)
vpa.cell.neg.raw %>% 
  select(Samples, Group, starts_with("VPAnC")) %>% 
  gather("Compound", value = "Abundance", -c(Samples, Group)) %>% 
  mutate(Cmpnd_sort = factor(Compound, levels = vpa.cell.neg.raw.grp.mean.order$Compound)) %>% 
  ggplot(aes(Cmpnd_sort, log2(Abundance), color = Group, group = Samples)) + 
  geom_line(alpha = 0.2, size = 1) + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab("Compound") +
  geom_line(
    data = vpa.cell.neg.raw.grp.mean %>% 
      mutate(Cmpnd_sort = factor(Compound, levels = vpa.cell.neg.raw.grp.mean.order$Compound)), 
    aes(Cmpnd_sort, log2(Group_mean_abun), color = Group, group = Group),
    size = 0.5
    ) +
  ggtitle("Profile Plot of all compound abundances\nWith average per sample Group overlaid\nVPA Dose Exp / Cells / Negative Mode")
vpa.cell.neg.raw.grp.mean %>% 
  mutate(Cmpnd_sort = factor(Compound, levels = vpa.cell.neg.raw.grp.mean.order$Compound)) %>% 
  ggplot(aes(Cmpnd_sort, log2(Group_mean_abun), color = Group, group = Group)) +
  geom_point(size = 1, alpha = 0.8) +
  geom_line(alpha = 0.8) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab("Compound") +
  ylab("log2(Sample Group Mean)") +
  ggtitle("Profile Plot of compound means by sample Group only\nVPA Dose Exp / Cells / Negative Mode")
vpa.cell.neg.raw.grp.diff <- vpa.cell.neg.raw.grp.mean %>% 
  spread(Group, Group_mean_abun) %>% 
  mutate(smpl_solv_diff = sample / solv)
vpa.cell.neg.raw.grp.diff %>% 
  ggplot(aes(log2(smpl_solv_diff))) +
  geom_histogram(bins = 50)
# include compounds with FC > 2.5 or FC is NA (indication of NA in solv)
vpa.cell.neg.cmpnd.to.incl <- vpa.cell.neg.raw.grp.diff %>% 
  filter(smpl_solv_diff > 2.5 | is.na(smpl_solv_diff))
# original number of metabolites
nrow(vpa.cell.neg.raw.grp.diff)
# number after filtering
nrow(vpa.cell.neg.cmpnd.to.incl)
```


### Cells / Positive Mode

```{r solv_cell_pos}
vpa.cell.pos.raw.grp.mean <- vpa.cell.pos.raw %>% 
  group_by(Group) %>% 
  summarize_at(vars(matches("VPApC")), mean, na.rm = TRUE) %>% 
  gather(key = "Compound", value = "Group_mean_abun", -Group)
vpa.cell.pos.raw.grp.mean %>% 
  ggplot(aes(log2(Group_mean_abun), color = Group)) +
  geom_density(size = 2, alpha = 0.8) +
  ggtitle("Distribution of compound means\nVPA Dose Exp / Cells / Positive Mode\nGrouped by sample Group")
vpa.cell.pos.raw.grp.mean.order <- vpa.cell.pos.raw.grp.mean %>% 
  filter(Group == "sample") %>% 
  arrange(Group_mean_abun)
vpa.cell.pos.raw %>% 
  select(Samples, Group, starts_with("VPApC")) %>% 
  gather("Compound", value = "Abundance", -c(Samples, Group)) %>% 
  mutate(Cmpnd_sort = factor(Compound, levels = vpa.cell.pos.raw.grp.mean.order$Compound)) %>% 
  ggplot(aes(Cmpnd_sort, log2(Abundance), color = Group, group = Samples)) + 
  geom_line(alpha = 0.2, size = 1) + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab("Compound") +
  geom_line(
    data = vpa.cell.pos.raw.grp.mean %>% 
      mutate(Cmpnd_sort = factor(Compound, levels = vpa.cell.pos.raw.grp.mean.order$Compound)), 
    aes(Cmpnd_sort, log2(Group_mean_abun), color = Group, group = Group),
    size = 0.5
    ) +
  ggtitle("Profile Plot of all compound abundances\nWith average per sample Group overlaid\nVPA Dose Exp / Cells / Positive Mode")
vpa.cell.pos.raw.grp.mean %>% 
  mutate(Cmpnd_sort = factor(Compound, levels = vpa.cell.pos.raw.grp.mean.order$Compound)) %>% 
  ggplot(aes(Cmpnd_sort, log2(Group_mean_abun), color = Group, group = Group)) +
  geom_point(size = 1, alpha = 0.8) +
  geom_line(alpha = 0.8) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab("Compound") +
  ylab("log2(Sample Group Mean)") +
  ggtitle("Profile Plot of compound means by sample Group only\nVPA Dose Exp / Cells / Positive Mode")
vpa.cell.pos.raw.grp.diff <- vpa.cell.pos.raw.grp.mean %>% 
  spread(Group, Group_mean_abun) %>% 
  mutate(smpl_solv_diff = sample / solv)
vpa.cell.pos.raw.grp.diff %>% 
  ggplot(aes(log2(smpl_solv_diff))) +
  geom_histogram(bins = 50)
# include compounds with FC > 2.5 or FC is NA (indication of NA in solv)
vpa.cell.pos.cmpnd.to.incl <- vpa.cell.pos.raw.grp.diff %>% 
  filter(smpl_solv_diff > 2.5 | is.na(smpl_solv_diff)) %>% 
  filter(!(Compound %in% vpa.cell.pos.cmpnd.excl$Compound))
# original number
nrow(vpa.cell.pos.raw.grp.diff)
# number after filtering
nrow(vpa.cell.pos.cmpnd.to.incl)
```


# Data Prep and Preliminary Analysis

## Cleanup

```{r min_na_rpl_log_transform, comment=NA}
vpa.cell.neg.noNA <- vpa.cell.neg.raw %>% 
  select(Samples:Dose_uM, one_of(vpa.cell.neg.cmpnd.to.incl$Compound)) %>% 
  ReplaceNAwMinLogTransformSingle("VPAnC")
vpa.cell.pos.noNA <- vpa.cell.pos.raw %>% 
  select(Samples:Dose_uM, one_of(vpa.cell.pos.cmpnd.to.incl$Compound)) %>% 
  ReplaceNAwMinLogTransformSingle("VPApC")
```


## Distribution plots

### Cells / Negative Mode

```{r distribution_plots_cells_neg, comment=NA}
vpa.cell.neg.noNA.gathered <- vpa.cell.neg.noNA %>% 
  gather(
    key = "Metabolite", "Abundance", 
    which(colnames(vpa.cell.neg.noNA) == "VPAnC1"):ncol(vpa.cell.neg.noNA)
    )
vpa.cell.neg.noNA.gathered %>% 
  ggplot(aes(Samples, Abundance, fill = Group)) +
  geom_boxplot() +
  geom_boxplot(aes(color = Group), fatten = NULL, fill = NA, coef = 0, outlier.alpha = 0, show.legend = FALSE) +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab("log2(Abundance)") +
  ggtitle("Boxplot of compound abundances\nAll samples\nVPA Dose Exp / Cells / Negative Mode")
# same data format, but as ridge plots
vpa.cell.neg.noNA.gathered %>% 
  ggplot(aes(y = Samples, x = Abundance, fill = Group)) + 
  geom_density_ridges(scale = 15) +
  theme_ridges() +
  scale_y_discrete(expand = c(0.01, 0)) +
  ggtitle("Ridge plot of compound abundances\nAll samples\nVPA Dose Exp / Cells / Negative Mode")
# experimental samples only
vpa.cell.neg.noNA.gathered %>% 
  filter(Group == "sample") %>% 
  ggplot(aes(y = Samples, x = Abundance, fill = Dose_uM)) + 
  geom_density_ridges(scale = 10) +
  theme_ridges() +
  scale_y_discrete(expand = c(0.01, 0)) +
  ggtitle("Ridge plot of compound abundances\nExperimental samples only\nVPA Dose Exp / Cells / Negative Mode")
# overlay the distributions for another look
vpa.cell.neg.noNA.gathered %>%
  filter(Group == "sample") %>% 
  ggplot(aes(Abundance, group = Samples, color = Dose_uM)) +
  geom_density(alpha = 0.8, size = 0.75) +
  xlab("log2(Abundance)") +
  ggtitle("Density plot of compound abundances\nExperimental samples only\nVPA Dose Exp / Cells / Negative Mode")
```


### Cells / Positive Mode

```{r distribution_plots_cells_pos, comment=NA}
vpa.cell.pos.noNA.gathered <- vpa.cell.pos.noNA %>% 
  gather(
    key = "Metabolite", "Abundance", 
    which(colnames(vpa.cell.pos.noNA) == "VPApC1"):ncol(vpa.cell.pos.noNA)
    )
vpa.cell.pos.noNA.gathered %>% 
  ggplot(aes(Samples, Abundance, fill = Group)) +
  geom_boxplot() +
  geom_boxplot(aes(color = Group), fatten = NULL, fill = NA, coef = 0, outlier.alpha = 0, show.legend = FALSE) +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab("log2(Abundance)") +
  ggtitle("Boxplot of compound abundances\nAll samples\nVPA Dose Exp / Cells / Positive Mode")
vpa.cell.pos.noNA.gathered %>% 
  ggplot(aes(y = Samples, x = Abundance, fill = Group)) + 
  geom_density_ridges(scale = 15) +
  theme_ridges() +
  scale_y_discrete(expand = c(0.01, 0)) +
  ggtitle("Ridge plot of compound abundances\nAll samples\nVPA Dose Exp / Cells / Positive Mode")
vpa.cell.pos.noNA.gathered %>% 
  filter(Group == "sample") %>% 
  ggplot(aes(y = Samples, x = Abundance, fill = Dose_uM)) + 
  geom_density_ridges(scale = 10) +
  theme_ridges() +
  scale_y_discrete(expand = c(0.01, 0)) +
  ggtitle("Ridge plot of compound abundances\nExperimental samples only\nVPA Dose Exp / Cells / Positive Mode")
vpa.cell.pos.noNA.gathered %>%
  filter(Group == "sample") %>% 
  ggplot(aes(Abundance, group = Samples, color = Dose_uM)) +
  geom_density(alpha = 0.8, size = 0.75) +
  xlab("log2(Abundance)") +
  ggtitle("Density plot of compound abundances\nExperimental samples only\nVPA Dose Exp / Cells / Positive Mode")
```


## Principal Component Analysis 

Some plots to understand the relationship between the samples, QC samples, solvent, and empty samples in some cases. 

### Cells / Negative Mode

```{r pca_cells_neg, comment=NA}
### PCA on all Samples ###
vpa.cell.neg.full.pca <- vpa.cell.neg.noNA %>% 
  select(starts_with("VPAnC")) %>% 
  # good idea to center data before pca, but scaling should not be necessary
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
# plot variance explained by each new principal component
plot(
  (vpa.cell.neg.full.pca$sdev ^ 2) * 100 / sum(vpa.cell.neg.full.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nAll samples only\nVPA Dose Exp / Cells / Negative Mode",
  type = "b"
  )
vpa.cell.neg.full.pca.x <- as.data.frame(vpa.cell.neg.full.pca$x)
row.names(vpa.cell.neg.full.pca.x) <- vpa.cell.neg.noNA$Samples
vpa.cell.neg.full.pca.x <- vpa.cell.neg.full.pca.x %>% 
  bind_cols(vpa.cell.neg.noNA %>% select(Group:Dose_uM))
vpa.cell.neg.full.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (97.3% Var)") +
  ylab("PC2 (1.0%)") +
  ggtitle("Principal Component Analysis\nAll Samples\nVPA Dose Exp / Cells / Negative Mode")

### Experimental Samples Only ###
vpa.cell.neg.smpl.pca <- vpa.cell.neg.noNA %>% 
  filter(Group == "sample") %>% 
  select(starts_with("VPAnC")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (vpa.cell.neg.smpl.pca$sdev ^ 2) * 100 / sum(vpa.cell.neg.smpl.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nExperimental samples only\nVPA Dose Exp / Cells / Negative Mode",
  type = "b"
  )
vpa.cell.neg.smpl.pca.x <- as.data.frame(vpa.cell.neg.smpl.pca$x)
vpa.cell.neg.smpl.pca.x <- vpa.cell.neg.smpl.pca.x %>% 
  bind_cols(
    vpa.cell.neg.noNA %>% 
      filter(Group == "sample") %>% 
      select(Samples, Group:Dose_uM)
  )
row.names(vpa.cell.neg.smpl.pca.x) <- vpa.cell.neg.smpl.pca.x$Samples
vpa.cell.neg.smpl.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, color = Dose_uM)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (55.9% Var)") +
  ylab("PC2 (21.1%)") +
  ggtitle("Principal Component Analysis\nExperimental samples only\nVPA Dose Exp / Cells / Negative Mode")
vpa.cell.neg.smpl.pca.x %>% 
  ggplot(aes(x = PC3, y = PC4, color = Dose_uM)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC3 (6.3% Var)") +
  ylab("PC4 (4.1%)") +
  ggtitle("Principal Component Analysis\nExperimental samples only\nVPA Dose Exp / Cells / Negative Mode")
```


### Cells / Positive Mode

```{r pca_cells_pos, comment=NA}
### PCA on all Samples ###
vpa.cell.pos.full.pca <- vpa.cell.pos.noNA %>% 
  select(starts_with("VPApC")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (vpa.cell.pos.full.pca$sdev ^ 2) * 100 / sum(vpa.cell.pos.full.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nAll samples only\nVPA Dose Exp / Cells / Positive Mode",
  type = "b"
  )
vpa.cell.pos.full.pca.x <- as.data.frame(vpa.cell.pos.full.pca$x)
row.names(vpa.cell.pos.full.pca.x) <- vpa.cell.pos.noNA$Samples
vpa.cell.pos.full.pca.x <- vpa.cell.pos.full.pca.x %>% 
  bind_cols(vpa.cell.pos.noNA %>% select(Group:Dose_uM))
vpa.cell.pos.full.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (94.8% Var)") +
  ylab("PC2 (3.3%)") +
  ggtitle("Principal Component Analysis\nAll Samples\nVPA Dose Exp / Cells / Positive Mode")

### Experimental Samples Only ###
vpa.cell.pos.smpl.pca <- vpa.cell.pos.noNA %>% 
  filter(Group == "sample") %>% 
  select(starts_with("VPApC")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (vpa.cell.pos.smpl.pca$sdev ^ 2) * 100 / sum(vpa.cell.pos.smpl.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nExperimental samples only\nVPA Dose Exp / Cells / Positive Mode",
  type = "b"
  )
vpa.cell.pos.smpl.pca.x <- as.data.frame(vpa.cell.pos.smpl.pca$x)
vpa.cell.pos.smpl.pca.x <- vpa.cell.pos.smpl.pca.x %>% 
  bind_cols(
    vpa.cell.pos.noNA %>% 
      filter(Group == "sample") %>% 
      select(Samples, Group:Dose_uM)
  )
row.names(vpa.cell.pos.smpl.pca.x) <- vpa.cell.pos.smpl.pca.x$Samples
vpa.cell.pos.smpl.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, color = Dose_uM)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (51.1% Var)") +
  ylab("PC2 (24.6%)") +
  ggtitle("Principal Component Analysis\nExperimental samples only\nVPA Dose Exp / Cells / Positive Mode")
vpa.cell.pos.smpl.pca.x %>% 
  ggplot(aes(x = PC3, y = PC4, color = Dose_uM)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC3 (9.2% Var)") +
  ylab("PC4 (3.6%)") +
  ggtitle("Principal Component Analysis\nExperimental samples only\nVPA Dose Exp / Cells / Positive Mode")
```


# Statistical analysis

## Cells / Negative Mode

```{r vpa_cell_neg_stat, comment=NA}
# sample prep
vpa.cell.neg.smpl.data <- vpa.cell.neg.noNA %>% 
    filter(Group == "sample")
vpa.cell.neg.data.for.sva <- as.matrix(
  vpa.cell.neg.smpl.data[, which(
    colnames(vpa.cell.neg.smpl.data) == "VPAnC1"
    ):ncol(vpa.cell.neg.smpl.data)]
  )
row.names(vpa.cell.neg.data.for.sva) <- vpa.cell.neg.smpl.data$Samples
vpa.cell.neg.data.for.sva <- t(vpa.cell.neg.data.for.sva)
# pheno prep
vpa.cell.neg.data.pheno <- as.data.frame(vpa.cell.neg.smpl.data[, 5:6])
row.names(vpa.cell.neg.data.pheno) <- vpa.cell.neg.smpl.data$Samples
# sva calculation
vpa.cell.neg.mod.vpa <- model.matrix(~ as.factor(Treatment), data = vpa.cell.neg.data.pheno)
vpa.cell.neg.mod0 <- model.matrix(~ 1, data = vpa.cell.neg.data.pheno)
set.seed(2018)
num.sv(vpa.cell.neg.data.for.sva, vpa.cell.neg.mod.vpa, method = "be")
set.seed(2018)
num.sv(vpa.cell.neg.data.for.sva, vpa.cell.neg.mod.vpa, method = "leek")
set.seed(2018)
vpa.cell.neg.sv <- sva(vpa.cell.neg.data.for.sva, vpa.cell.neg.mod.vpa, vpa.cell.neg.mod0)

# extract the surrogate variables
vpa.cell.neg.surr.var <- as.data.frame(vpa.cell.neg.sv$sv)
colnames(vpa.cell.neg.surr.var) <- c("S1", "S2")
colnames(vpa.cell.neg.mod.vpa) <- c("cntrl", "VPAvsCNTRL")
# combine the full model matrix and the surrogate variables into one
vpa.cell.neg.d.sv <- cbind(vpa.cell.neg.mod.vpa, vpa.cell.neg.surr.var[, 2])
plot( vpa.cell.neg.data.pheno$Dose_uM, vpa.cell.neg.surr.var$S1)
plot( vpa.cell.neg.data.pheno$Dose_uM, vpa.cell.neg.surr.var$S2)
colnames(vpa.cell.neg.d.sv)[3] <- "S2"
vpa.cell.neg.top.table <- vpa.cell.neg.data.for.sva %>% 
  lmFit(vpa.cell.neg.d.sv) %>% 
  eBayes() %>% 
  topTable(coef = "VPAvsCNTRL", adjust = "bonferroni", p.value = 0.05, n = nrow(vpa.cell.neg.data.for.sva))
vpa.cell.neg.top.table
```  
 

## Cells / Positive Mode

```{r vpa_cell_pos_stat, comment=NA}
# sample prep
vpa.cell.pos.smpl.data <- vpa.cell.pos.noNA %>% 
    filter(Group == "sample")
vpa.cell.pos.data.for.sva <- as.matrix(
  vpa.cell.pos.smpl.data[, which(
    colnames(vpa.cell.pos.smpl.data) == "VPApC1"
    ):ncol(vpa.cell.pos.smpl.data)]
  )
row.names(vpa.cell.pos.data.for.sva) <- vpa.cell.pos.smpl.data$Samples
vpa.cell.pos.data.for.sva <- t(vpa.cell.pos.data.for.sva)
# pheno prep
vpa.cell.pos.data.pheno <- as.data.frame(vpa.cell.pos.smpl.data[, 5:6])
row.names(vpa.cell.pos.data.pheno) <- vpa.cell.pos.smpl.data$Samples
# sva calculation
vpa.cell.pos.mod.vpa <- model.matrix(~ as.factor(Treatment), data = vpa.cell.pos.data.pheno)
vpa.cell.pos.mod0 <- model.matrix(~ 1, data = vpa.cell.pos.data.pheno)
set.seed(2018)
num.sv(vpa.cell.pos.data.for.sva, vpa.cell.pos.mod.vpa, method = "be")
set.seed(2018)
num.sv(vpa.cell.pos.data.for.sva, vpa.cell.pos.mod.vpa, method = "leek")
set.seed(2018)
vpa.cell.pos.sv <- sva(vpa.cell.pos.data.for.sva, vpa.cell.pos.mod.vpa, vpa.cell.pos.mod0)

# extract the surrogate variables
vpa.cell.pos.surr.var <- as.data.frame(vpa.cell.pos.sv$sv)
colnames(vpa.cell.pos.surr.var) <- "S1"
colnames(vpa.cell.pos.mod.vpa) <- c("cntrl", "VPAvsCNTRL")
# combine the full model matrix and the surrogate variables into one
plot(vpa.cell.pos.data.pheno$Dose_uM, vpa.cell.pos.surr.var$S1)
vpa.cell.pos.d.sv <- cbind(vpa.cell.pos.mod.vpa, vpa.cell.pos.surr.var)
vpa.cell.pos.top.table <- vpa.cell.pos.data.for.sva %>% 
  lmFit(vpa.cell.pos.d.sv) %>% 
  eBayes() %>% 
  topTable(coef = "VPAvsCNTRL", adjust = "bonferroni", p.value = 0.05, n = nrow(vpa.cell.pos.data.for.sva))
vpa.cell.pos.top.table
vpa.cell.pos.compound.info %>% filter(compound_short == "VPApC13")
```  
  

# Session Info

```{r session_info, comment=NA}
sessionInfo()
```

