---
title: "Lrp6 KO Exp 2"
author: "Darya Akimova"
date: "January 2019"
output: 
  html_document:
    toc: true
    toc_float:
      smooth_scroll: false
    number_sections: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Setup

## Packages

```{r packages, comment=NA, message = FALSE}
library(tidyverse)
library(cowplot)
library(GGally)
library(heatmaply)
library(sva)
library(limma)
library(biobroom)
library(ggridges)
theme_set(theme_minimal())
```


## Data

### Metabolite Abundances

```{r abundance_import, comment=NA}
ko2.neg.raw <- read_csv("./data/abundances/lrp6ko_exp2_abundances_negmode.csv")
ko2.pos.raw <- read_csv("./data/abundances/lrp6ko_exp2_abundances_posmode.csv")
setequal(ko2.neg.raw$Samples, ko2.pos.raw$Samples)
```


### Compound Information

```{r compound_info_import, comment=NA}
ko2.neg.compound.info <- read_csv("./data/compound_info/lrp6ko_exp2_compound_info_negmode.csv")
ko2.pos.compound.info <- read_csv("./data/compound_info/lrp6ko_exp2_compound_info_posmode.csv")
```


## Functions

```{r functions, comment=NA}
MissingPerSamplePlot <- function(raw.data, start.string) {
  # Counts the number of missing/NA values per sample and
  # percent compounds missing out of total number of compounds per sample
  # Then passes the result into a vertical bar plot, where each 
  # bar represents a single sample and the size of the bar 
  # is the % of compounds missing
  
  counted.na <- raw.data %>%
  select(starts_with(start.string)) %>% 
  mutate(
    count.na = apply(., 1, function(x) sum(is.na(x))),
    percent.na = (count.na / ncol(raw.data %>% select(starts_with(start.string)))) * 100
    ) %>%
  dplyr::select(count.na, percent.na) %>%
  bind_cols(
    raw.data %>% 
      select(Samples, Type)
      ) %>% 
  arrange(percent.na) %>% 
  mutate(f.order = factor(Samples, levels = Samples))
counted.na %>% 
  ggplot(aes(x = f.order, y = percent.na, fill = Type)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 20, color = "gray", size = 1, alpha = 0.8) +
  coord_flip()+
  xlab("Samples") +
  ylab("Percent missing values in sample") +
  theme(axis.text.y = element_text(size = 6)) 
}
MissingPerCompound <- function(raw.data, start.string){
  # Function to count in how many experimental samples each compound is missing
  # Also calculates the % missing:
  # (# NA per compound across all experimental samples) * 100 / (tot num of samples)
  
  raw.data %>% 
  filter(Type == "sample") %>% 
  select(Samples, starts_with(start.string)) %>% 
  gather(key = "Compound", value = "Abundance", -Samples) %>% 
  group_by(Compound) %>% 
  summarise(
    na_count = sum(is.na(Abundance)),
    n_samples = n(),
    percent_na = (na_count * 100 / n_samples)
    ) %>% 
  filter(na_count > 0) %>% 
  arrange(desc(percent_na))
}
ReplaceNAwMinLogTransform <- function(raw.dataframe, start.prefix, type) {
  # Function to replace any NAs in each column with the minimum for that column, 
  # separately for each data.set (patient and dabase samples separately)
  #
  # Returns: returns data frame(? check on this) with only the values bound by the function args
  # 
  smpls <- raw.dataframe %>%
    filter(Type == "sample") %>% 
    dplyr::select(starts_with(start.prefix))
  smpls.min <- lapply(smpls, min, na.rm = TRUE)
  # replace the missing values in the real samples with the minimum of the samples
  # then take the log
  smpls.noNA <- raw.dataframe  %>%
    filter(Type == "sample") %>% 
    dplyr::select(Samples:Status) %>%
    bind_cols(
      smpls %>%
        replace_na(replace = smpls.min) %>% 
        log2()
      ) 
  # replace missing QC values with values within the QC groups
  QC <- raw.dataframe %>%
    filter(Type == "QC") %>% 
    dplyr::select(starts_with(start.prefix))
  QC.min <- lapply(QC, min, na.rm = TRUE)
  QC.noNA <- raw.dataframe  %>%
    filter(Type == "QC") %>% 
    dplyr::select(Samples:Status) %>%
    bind_cols(
      QC %>%
        replace_na(replace = QC.min) %>% 
        log2()
      )
  # replace the missing values in solv and empty samples with 1 - for PCA analysis
  # then take the log
  other.min <- setNames(
    as.list(
      rep(2, ncol(
        raw.dataframe %>% 
          dplyr::select(starts_with(start.prefix))))
      ),
    colnames(raw.dataframe %>% dplyr::select(starts_with(start.prefix)))
                )
  other.num.log <- raw.dataframe %>%
    filter(Type == "Solv") %>% 
    dplyr::select(Samples:Status) %>%
    bind_cols(
      raw.dataframe %>% 
        filter(Type == "Solv") %>% 
        dplyr::select(starts_with(start.prefix)) %>%
        replace_na(replace = other.min) %>% 
        log2()
      )
  # combine them together back into one data frame
  all.noNA <- smpls.noNA %>% 
    bind_rows(QC.noNA) %>% 
    bind_rows(other.num.log)
}
HeatmapPrepAlt <- function(raw.data, start.prefix){
  # function for preparing dara for heatmap viz
  x <- raw.data %>% 
    select(starts_with(start.prefix)) %>% 
    scale(center = TRUE, scale = TRUE) %>% 
    as.matrix() 
  row.names(x) <- raw.data$Samples
  return(x)
}
```


# Data Exploration

## Mass vs Retention Time

Q: What are the distributions of compound masses and retention times?

```{r mass_v_RT_plots, fig.height=7, fig.width=9, comment=NA}
full.ko2.cmpnd <- ko2.neg.compound.info %>% 
  mutate(Set = "Neg") %>% 
  bind_rows(ko2.pos.compound.info %>% mutate(Set = "Pos"))
full.ko2.cmpnd %>% 
  ggplot(aes(x = RT, y = Mass)) +
  geom_point(size = 3, alpha = 0.3) +
  xlab("Retention Time (min)") +
  ylab("Mass (Da)") +
  ggtitle("Mass v RT\nVPA + FA Exp") +
  ylim(0, 1000)
full.ko2.cmpnd %>% 
  ggplot(aes(x = RT, y = Mass, color = Set)) +
  geom_point(size = 3, alpha = 0.8) +
  xlab("Retnetion Time (min)") +
  ylab("Mass (Da)") +
  ggtitle("Mass v RT\nVPA + FA Exp") +
  facet_wrap(~ Set) +
  ylim(0, 1000)
```


Q: Which compounds were found in one or more of the data types?


```{r cmnpd_mode_match, fig.height= 5, fig.width=8, comment=NA}
ko2.cmpnd.join <- ko2.neg.compound.info %>% 
  inner_join(ko2.pos.compound.info, by = "CAS_ID", suffix = c("_n", "_p")) %>% 
  select(
    CAS_ID, contains("short"), 
    contains("full"), starts_with("Formula"), 
    starts_with("Mass"), starts_with("RT")
    )
# compound names - found in pos and neg mode / cells 
print(ko2.cmpnd.join$Compound_full_n)
# percent of cell / neg compounds found in cell / pos 
round(nrow(ko2.cmpnd.join) * 100 / nrow(ko2.neg.compound.info), 1)
# percent of cell / neg compounds found in cell / pos 
round(nrow(ko2.cmpnd.join) * 100 / nrow(ko2.pos.compound.info), 1)
# problem checks #
ko2.cmpnd.join %>% 
  select(contains("Mass")) %>% 
  ggpairs()
ko2.cmpnd.join %>% 
  select(starts_with("RT")) %>% 
  ggpairs()
```


## Missing Values

Q: Do any of the samples have greater than 20% missing (NA) compound abundances, out of all of the features in the dataset?

```{r missing_per_sample_plots, comment=NA}
MissingPerSamplePlot(ko2.neg.raw, "Lrp6KO2n") +
  ggtitle("Missing Per Sample\nLrp6KO / Exp2 / Neg Mode")
MissingPerSamplePlot(ko2.pos.raw, "Lrp6KO2p") +
  ggtitle("Missing Per Sample\nLrp6KO / Exp2 / Pos Mode")
```

I think I had mixed up the QC and Solvent samples in this experiment when running. 

Q: Are any of the compounds more than 20% missing in the experimental sample group? If there are any, they will be excluded from analysis. 

```{r percent_missing_per_compound, comment=NA}
(ko2.neg.cmpnd.to.excl <- MissingPerCompound(ko2.neg.raw, "Lrp6KO2n") %>% 
  filter(percent_na > 20))
(ko2.pos.cmpnd.to.excl <- MissingPerCompound(ko2.pos.raw, "Lrp6KO2p") %>% 
  filter(percent_na > 20))
```


## Negative Controls and Compound Elimination

### Negative Mode

```{r empty_solv_neg, comment=NA}
ko2.neg.raw.type.mean <- ko2.neg.raw %>% 
  group_by(Type) %>% 
  summarize_at(vars(matches("Lrp6KO2n")), mean, na.rm = TRUE) %>% 
  gather(key = "Compound", value = "Type_mean_abun", -Type)
ko2.neg.raw.type.mean %>% 
  ggplot(aes(log2(Type_mean_abun), color = Type)) +
  geom_density(size = 2, alpha = 0.8) +
  ggtitle("Distribution of compound means\nLrp6KO Exp 2 / Negative Mode\nGrouped by sample type")
ko2.neg.raw.type.mean.order <- ko2.neg.raw.type.mean %>% 
  filter(Type == "sample") %>% 
  arrange(Type_mean_abun)
ko2.neg.raw %>% 
  select(Samples, Type, starts_with("Lrp6KO2n")) %>% 
  gather("Compound", value = "Abundance", -c(Samples, Type)) %>% 
  mutate(Cmpnd_sort = factor(Compound, levels = ko2.neg.raw.type.mean.order$Compound)) %>% 
  ggplot(aes(Cmpnd_sort, log2(Abundance), color = Type, group = Samples)) + 
  geom_line(alpha = 0.1, size = 1) + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab("Compound") +
  geom_line(
    data = ko2.neg.raw.type.mean %>% 
      mutate(Cmpnd_sort = factor(Compound, levels = ko2.neg.raw.type.mean.order$Compound)), 
    aes(Cmpnd_sort, log2(Type_mean_abun), color = Type, group = Type),
    size = 0.5
    ) +
  ggtitle("Profile Plot of all compound abundances\nWith average per sample type overlaid\nLrp6KO Exp 2 / Negative Mode")
ko2.neg.raw.type.mean %>% 
  mutate(Cmpnd_sort = factor(Compound, levels = ko2.neg.raw.type.mean.order$Compound)) %>% 
  ggplot(aes(Cmpnd_sort, log2(Type_mean_abun), color = Type, group = Type)) +
  geom_point(size = 1, alpha = 0.8) +
  geom_line(alpha = 0.8) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab("Compound") +
  ylab("log2(Sample Type Mean)") +
  ggtitle("Profile Plot of compound means by sample type only\nLrp6KO Exp 2 / Negative Mode")
ko2.neg.raw.type.diff <- ko2.neg.raw.type.mean %>% 
  spread(Type, Type_mean_abun) %>% 
  mutate(smpl_solv_diff = sample / Solv)
ko2.neg.raw.type.diff %>% 
  ggplot(aes(log2(smpl_solv_diff))) +
  geom_histogram(bins = 50)
# include compounds with FC > 2.5 or FC is NA (indication of NA in solv or empty)
ko2.neg.cmpnd.to.incl <- ko2.neg.raw.type.diff  %>% 
  filter(smpl_solv_diff > 2.5 | is.na(smpl_solv_diff)) %>% 
  filter(!(Compound %in% ko2.neg.cmpnd.to.excl$Compound))
# original number of metabolites
nrow(ko2.neg.raw.type.diff)
# number of metabolites after filtering 
nrow(ko2.neg.cmpnd.to.incl)
```


### Positive Mode

```{r empty_solv_pos, comment=NA}
ko2.pos.raw.type.mean <- ko2.pos.raw %>% 
  group_by(Type) %>% 
  summarize_at(vars(matches("Lrp6KO2p")), mean, na.rm = TRUE) %>% 
  gather(key = "Compound", value = "Type_mean_abun", -Type)
ko2.pos.raw.type.mean %>% 
  ggplot(aes(log2(Type_mean_abun), color = Type)) +
  geom_density(size = 2, alpha = 0.8) +
  ggtitle("Distribution of compound means\nLrp6KO Exp 2 / Positive Mode\nGrouped by sample type")
ko2.pos.raw.type.mean.order <- ko2.pos.raw.type.mean %>% 
  filter(Type == "sample") %>% 
  arrange(Type_mean_abun)
ko2.pos.raw %>% 
  select(Samples, Type, starts_with("Lrp6KO2p")) %>% 
  gather("Compound", value = "Abundance", -c(Samples, Type)) %>% 
  mutate(Cmpnd_sort = factor(Compound, levels = ko2.pos.raw.type.mean.order$Compound)) %>% 
  ggplot(aes(Cmpnd_sort, log2(Abundance), color = Type, group = Samples)) + 
  geom_line(alpha = 0.1, size = 1) + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab("Compound") +
  geom_line(
    data = ko2.pos.raw.type.mean %>% 
      mutate(Cmpnd_sort = factor(Compound, levels = ko2.pos.raw.type.mean.order$Compound)), 
    aes(Cmpnd_sort, log2(Type_mean_abun), color = Type, group = Type),
    size = 0.5
    ) +
  ggtitle("Profile Plot of all compound abundances\nWith average per sample type overlaid\nLrp6KO Exp 2 / Positive Mode")
ko2.pos.raw.type.mean %>% 
  mutate(Cmpnd_sort = factor(Compound, levels = ko2.pos.raw.type.mean.order$Compound)) %>% 
  ggplot(aes(Cmpnd_sort, log2(Type_mean_abun), color = Type, group = Type)) +
  geom_point(size = 1, alpha = 0.8) +
  geom_line(alpha = 0.8) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab("Compound") +
  ylab("log2(Sample Type Mean)") +
  ggtitle("Profile Plot of compound means by sample type only\nLrp6KO Exp 2 / Positive Mode")
ko2.pos.raw.type.diff <- ko2.pos.raw.type.mean %>% 
  spread(Type, Type_mean_abun) %>% 
  mutate(smpl_solv_diff = sample / Solv)
ko2.pos.raw.type.diff %>% 
  ggplot(aes(log2(smpl_solv_diff))) +
  geom_histogram(bins = 50)
# include compounds with FC > 2.5 or FC is NA (indication of NA in solv or empty)
ko2.pos.cmpnd.to.incl <- ko2.pos.raw.type.diff  %>% 
  filter(smpl_solv_diff > 2.5 | is.na(smpl_solv_diff)) %>% 
  filter(!(Compound %in% ko2.pos.cmpnd.to.excl$Compound))
# original number of metabolites
nrow(ko2.pos.raw.type.diff)
# number of metabolites after filtering 
nrow(ko2.pos.cmpnd.to.incl)
```


# Data Prep and Preliminary Analysis

## Cleanup

```{r min_na_rpl_log_transform, comment=NA}
ko2.neg.noNA <- ko2.neg.raw %>% 
  select(Samples:Status, one_of(ko2.neg.cmpnd.to.incl$Compound)) %>% 
  ReplaceNAwMinLogTransform("Lrp6KO2n")
ko2.pos.noNA <- ko2.pos.raw %>% 
  select(Samples:Status, one_of(ko2.pos.cmpnd.to.incl$Compound)) %>% 
  ReplaceNAwMinLogTransform("Lrp6KO2p")
```


## Distribution plots

### Negative Mode

```{r distribution_plots_cells_neg, comment=NA}
vf.cell.neg.noNA.gathered <- vf.cell.neg.noNA %>% 
  gather(
    key = "Metabolite", "Abundance", 
    which(colnames(vf.cell.neg.noNA) == "VPA_FAnC10"):ncol(vf.cell.neg.noNA)
    )
# plot all abundances in a sample, grouped by sample as a boxplot
vf.cell.neg.noNA.gathered %>% 
  ggplot(aes(Samples, Abundance, fill = Group)) +
  geom_boxplot() +
  geom_boxplot(aes(color = Group), fatten = NULL, fill = NA, coef = 0, outlier.alpha = 0, show.legend = FALSE) +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab("log2(Abundance)") +
  ggtitle("Boxplot of compound abundances\nAll samples\nVPA + FA / Cells / Negative Mode")
# same data format, but as ridge plots
vf.cell.neg.noNA.gathered %>% 
  ggplot(aes(y = Samples, x = Abundance, fill = Group)) + 
  geom_density_ridges(scale = 15) +
  theme_ridges() +
  scale_y_discrete(expand = c(0.01, 0)) +
  ggtitle("Ridge plot of compound abundances\nAll samples\nVPA + FA / Cells / Negative Mode")
# experimental samples only
vf.cell.neg.noNA.gathered %>% 
  filter(Group == "sample") %>% 
  unite("Treatment", VPA:FA, sep = "_") %>% 
  ggplot(aes(y = Samples, x = Abundance, fill = Treatment)) + 
  geom_density_ridges(scale = 10) +
  theme_ridges() +
  scale_y_discrete(expand = c(0.01, 0)) +
  ggtitle("Ridge plot of compound abundances\nExperimental samples only\nVPA + FA / Cells / Negative Mode")
# overlay the distributions for another look
vf.cell.neg.noNA.gathered %>%
  filter(Group == "sample") %>% 
  unite("Treatment", VPA:FA, sep = "_") %>% 
  ggplot(aes(Abundance, group = Samples, color = Treatment)) +
  geom_density(alpha = 0.8, size = 0.75) +
  xlab("log2(Abundance)") +
  ggtitle("Density plot of compound abundances\nExperimental samples only\nVPA + FA / Cells / Negative Mode")
```


### Positive Mode

```{r distribution_plots_cells_pos, comment=NA}
vf.cell.pos.noNA.gathered <- vf.cell.pos.noNA %>% 
  gather(
    key = "Metabolite", "Abundance", 
    which(colnames(vf.cell.pos.noNA) == "VPA_FApC10"):ncol(vf.cell.pos.noNA)
    )
vf.cell.pos.noNA.gathered %>% 
  ggplot(aes(Samples, Abundance, fill = Group)) +
  geom_boxplot() +
  geom_boxplot(aes(color = Group), fatten = NULL, fill = NA, coef = 0, outlier.alpha = 0, show.legend = FALSE) +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab("log2(Abundance)") +
  ggtitle("Boxplot of compound abundances\nAll samples\nVPA + FA / Cells / Positive Mode")
vf.cell.pos.noNA.gathered %>% 
  ggplot(aes(y = Samples, x = Abundance, fill = Group)) + 
  geom_density_ridges(scale = 15) +
  theme_ridges() +
  scale_y_discrete(expand = c(0.01, 0)) +
  ggtitle("Ridge plot of compound abundances\nAll samples\nVPA + FA / Cells / Positive Mode")
vf.cell.pos.noNA.gathered %>% 
  filter(Group == "sample") %>% 
  unite("Treatment", VPA:FA, sep = "_") %>% 
  ggplot(aes(y = Samples, x = Abundance, fill = Treatment)) + 
  geom_density_ridges(scale = 10) +
  theme_ridges() +
  scale_y_discrete(expand = c(0.01, 0)) +
  ggtitle("Ridge plot of compound abundances\nExperimental samples only\nVPA + FA / Cells / Positive Mode")
vf.cell.pos.noNA.gathered %>%
  filter(Group == "sample") %>% 
  unite("Treatment", VPA:FA, sep = "_") %>% 
  ggplot(aes(Abundance, group = Samples, color = Treatment)) +
  geom_density(alpha = 0.8, size = 0.75) +
  xlab("log2(Abundance)") +
  ggtitle("Density plot of compound abundances\nExperimental samples only\nVPA + FA / Cells / Positive Mode")
```


## Principal Component Analysis 

Some plots to understand the relationship between the samples, QC samples, solvent, and empty samples in some cases. 

### Negative Mode 

```{r pca_cells_neg, comment=NA}
### PCA on all Samples ###
vf.cell.neg.full.pca <- vf.cell.neg.noNA %>% 
  select(starts_with("VPA_FAnC")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (vf.cell.neg.full.pca$sdev ^ 2) * 100 / sum(vf.cell.neg.full.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nAll samples only\nVPA + FA / Cells / Negative Mode",
  type = "b"
  )
vf.cell.neg.full.pca.x <- as.data.frame(vf.cell.neg.full.pca$x)
row.names(vf.cell.neg.full.pca.x) <- vf.cell.neg.noNA$Samples
vf.cell.neg.full.pca.x <- vf.cell.neg.full.pca.x %>% 
  bind_cols(vf.cell.neg.noNA %>% select(Group:Plate))
vf.cell.neg.full.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (91.7% Var)") +
  ylab("PC2 (4.2% Var)") +
  ggtitle("Principal Component Analysis\nAll Samples\nVPA + FA / Cells / Negative Mode")

### Samples and QC ###
vf.cell.neg.smpl.QC.pca <- vf.cell.neg.noNA %>% 
  filter(Group == "sample" | Group == "QC") %>% 
  select(starts_with("VPA_FAnC")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (vf.cell.neg.smpl.QC.pca$sdev ^ 2) * 100 / sum(vf.cell.neg.smpl.QC.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nSamples and QC\nVPA + FA / Cells / Negative Mode",
  type = "b"
  )
vf.cell.neg.smpl.QC.pca.x <- as.data.frame(vf.cell.neg.smpl.QC.pca$x)
vf.cell.neg.smpl.QC.pca.x <- vf.cell.neg.smpl.QC.pca.x %>% 
  bind_cols(
    vf.cell.neg.noNA %>% 
      filter(Group == "sample" | Group == "QC") %>% 
      select(Samples, Group:Plate)
  )
row.names(vf.cell.neg.smpl.QC.pca.x) <- vf.cell.neg.smpl.QC.pca.x$Samples
vf.cell.neg.smpl.QC.pca.x %>% 
  unite("Treatment", VPA:FA, sep = "_") %>% 
  ggplot(aes(x = PC1, y = PC2, color = Treatment)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (32.3% Var)") +
  ylab("PC2 (25.1% Var)") +
  ggtitle("Principal Component Analysis\nSamples and QC\nVPA + FA / Cells / Negative Mode")
vf.cell.neg.smpl.QC.pca.x %>% 
  unite("Treatment", VPA:FA, sep = "_") %>% 
  ggplot(aes(x = PC3, y = PC4, color = Treatment)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC3 (13.7% Var)") +
  ylab("PC4 (6.5% Var)") +
  ggtitle("Principal Component Analysis\nSamples and QC\nVPA + FA / Cells / Negative Mode")

### Experimental Samples Only ###
vf.cell.neg.smpl.pca <- vf.cell.neg.noNA %>% 
  filter(Group == "sample") %>% 
  select(starts_with("VPA_FAnC")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (vf.cell.neg.smpl.pca$sdev ^ 2) * 100 / sum(vf.cell.neg.smpl.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nExperimental samples only\nVPA + FA / Cells / Negative Mode",
  type = "b"
  )
vf.cell.neg.smpl.pca.x <- as.data.frame(vf.cell.neg.smpl.pca$x)
vf.cell.neg.smpl.pca.x <- vf.cell.neg.smpl.pca.x %>% 
  bind_cols(
    vf.cell.neg.noNA %>% 
      filter(Group == "sample") %>% 
      select(Samples, Group:Plate)
  )
row.names(vf.cell.neg.smpl.pca.x) <- vf.cell.neg.smpl.pca.x$Samples
vf.cell.neg.smpl.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, color = FA, shape = VPA)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (44.5% Var)") +
  ylab("PC2 (19.3% Var)") +
  ggtitle("Principal Component Analysis\nExperimental samples only\nVPA + FA / Cells / Negative Mode")
vf.cell.neg.smpl.pca.x %>% 
  ggplot(aes(x = PC3, y = PC4, color = FA, shape = VPA)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC3 (10.4% Var)") +
  ylab("PC4 (7.7% Var)") +
  ggtitle("Principal Component Analysis\nExperimental samples only\nVPA + FA / Cells / Negative Mode")
```


### Positive Mode 

```{r pca_cells_pos, comment=NA}
### PCA on all Samples ###
vf.cell.pos.full.pca <- vf.cell.pos.noNA %>% 
  select(starts_with("VPA_FApC")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (vf.cell.pos.full.pca$sdev ^ 2) * 100 / sum(vf.cell.pos.full.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nAll samples only\nVPA + FA / Cells / Positive Mode",
  type = "b"
  )
vf.cell.pos.full.pca.x <- as.data.frame(vf.cell.pos.full.pca$x)
row.names(vf.cell.pos.full.pca.x) <- vf.cell.pos.noNA$Samples
vf.cell.pos.full.pca.x <- vf.cell.pos.full.pca.x %>% 
  bind_cols(vf.cell.pos.noNA %>% select(Group:Plate))
vf.cell.pos.full.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (92.0% Var)") +
  ylab("PC2 (2.7% Var)") +
  ggtitle("Principal Component Analysis\nAll Samples\nVPA + FA / Cells / Positive Mode")

### Samples and QC ###
vf.cell.pos.smpl.QC.pca <- vf.cell.pos.noNA %>% 
  filter(Group == "sample" | Group == "QC") %>% 
  select(starts_with("VPA_FApC")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (vf.cell.pos.smpl.QC.pca$sdev ^ 2) * 100 / sum(vf.cell.pos.smpl.QC.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nSamples and QC\nVPA + FA / Cells / Positive Mode",
  type = "b"
  )
vf.cell.pos.smpl.QC.pca.x <- as.data.frame(vf.cell.pos.smpl.QC.pca$x)
vf.cell.pos.smpl.QC.pca.x <- vf.cell.pos.smpl.QC.pca.x %>% 
  bind_cols(
    vf.cell.pos.noNA %>% 
      filter(Group == "sample" | Group == "QC") %>% 
      select(Samples, Group:Plate)
  )
row.names(vf.cell.pos.smpl.QC.pca.x) <- vf.cell.pos.smpl.QC.pca.x$Samples
vf.cell.pos.smpl.QC.pca.x %>% 
  unite("Treatment", VPA:FA, sep = "_", remove = FALSE) %>% 
  ggplot(aes(x = PC1, y = PC2, color = Treatment, shape = VPA)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (38.2% Var)") +
  ylab("PC2 (20.0% Var)") +
  ggtitle("Principal Component Analysis\nSamples and QC\nVPA + FA / Cells / Positive Mode")
vf.cell.pos.smpl.QC.pca.x %>% 
  unite("Treatment", VPA:FA, sep = "_") %>% 
  ggplot(aes(x = PC3, y = PC4, color = Treatment)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC3 (12.0% Var)") +
  ylab("PC4 (7.7% Var)") +
  ggtitle("Principal Component Analysis\nSamples and QC\nVPA + FA / Cells / Positive Mode")

### Experimental Samples Only ###
vf.cell.pos.smpl.pca <- vf.cell.pos.noNA %>% 
  filter(Group == "sample") %>% 
  select(starts_with("VPA_FApC")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (vf.cell.pos.smpl.pca$sdev ^ 2) * 100 / sum(vf.cell.pos.smpl.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nExperimental samples only\nVPA + FA / Cells / Positive Mode",
  type = "b"
  )
vf.cell.pos.smpl.pca.x <- as.data.frame(vf.cell.pos.smpl.pca$x)
vf.cell.pos.smpl.pca.x <- vf.cell.pos.smpl.pca.x %>% 
  bind_cols(
    vf.cell.pos.noNA %>% 
      filter(Group == "sample") %>% 
      select(Samples, Group:Plate)
  )
row.names(vf.cell.pos.smpl.pca.x) <- vf.cell.pos.smpl.pca.x$Samples
vf.cell.pos.smpl.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, color = FA, shape = VPA)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (49.4% Var)") +
  ylab("PC2 (15.7% Var)") +
  ggtitle("Principal Component Analysis\nExperimental samples only\nVPA + FA / Cells / Positive Mode")
vf.cell.pos.smpl.pca.x %>% 
  ggplot(aes(x = PC3, y = PC4, color = FA, shape = VPA)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC3 (9.8% Var)") +
  ylab("PC4 (7.6% Var)") +
  ggtitle("Principal Component Analysis\nExperimental samples only\nVPA + FA / Cells / Positive Mode")
```


# Batch Effects and Signifiance Testing

## Negative Mode

```{r vpa_fa_cells_neg_stat, comment=NA}
# sample prep
vf.cell.neg.smpl.data <- vf.cell.neg.noNA %>% 
  filter(Group == "sample") %>% 
  unite("Treatment", VPA:FA, sep = "_")
vf.cell.neg.data.for.sva <- as.matrix(
  vf.cell.neg.smpl.data[, which(
    colnames(vf.cell.neg.smpl.data) == "VPA_FAnC10"
    ):ncol(vf.cell.neg.smpl.data)]
  )
row.names(vf.cell.neg.data.for.sva) <- vf.cell.neg.smpl.data$Samples
vf.cell.neg.data.for.sva <- t(vf.cell.neg.data.for.sva)
# pheno prep
vf.cell.neg.data.pheno <- as.data.frame(vf.cell.neg.smpl.data[, 5:6])
row.names(vf.cell.neg.data.pheno) <- vf.cell.neg.smpl.data$Samples
# sva calculation
vf.cell.neg.mod.vf <- model.matrix(~ 0 + as.factor(Treatment), data = vf.cell.neg.data.pheno)
vf.cell.neg.mod0 <- model.matrix(~ 1, data = vf.cell.neg.data.pheno)
set.seed(2018)
num.sv(vf.cell.neg.data.for.sva, vf.cell.neg.mod.vf, method = "be")
set.seed(2018)
num.sv(vf.cell.neg.data.for.sva, vf.cell.neg.mod.vf, method = "leek")
set.seed(2018)
vf.cell.neg.sv <- sva(vf.cell.neg.data.for.sva, vf.cell.neg.mod.vf, vf.cell.neg.mod0)
# extract the surrogate variables
vf.cell.neg.surr.var <- as.data.frame(vf.cell.neg.sv$sv)
colnames(vf.cell.neg.surr.var) <- c("S1", "S2", "S3")

vf.cell.neg.covar <- vf.cell.neg.smpl.pca.x %>% 
  select(Samples, PC1:PC5) %>% 
  inner_join(
    cbind(vf.cell.neg.data.pheno, vf.cell.neg.surr.var) %>% 
      rownames_to_column("Samples"),
    by = "Samples"
    ) %>% 
  full_join(vf.cell.other.info, by = "Samples") %>% 
  as_tibble()
vf.cell.neg.covar %>% 
  select(Samples, Plate:S3) %>% 
  gather("surr_var", "value", S1:S3) %>% 
  ggplot(aes(Plate, value, color = Plate)) +
  geom_boxplot() +
  geom_jitter(size = 2, alpha = 0.8, width = 0.1) +
  facet_wrap(~ surr_var)
summary(aov(S3 ~ Plate, data = vf.cell.neg.covar))
vf.cell.neg.covar %>% 
  select(Samples, Treatment:S3) %>% 
  gather("surr_var", "value", S1:S3) %>% 
  ggplot(aes(Plate, value)) +
  geom_point(size = 3, alpha = 0.8, aes(color = Treatment)) +
  facet_wrap(~ surr_var)
vf.cell.neg.covar %>% 
  select(Samples:Plate) %>% 
  gather("PC", "value", PC1:PC5) %>% 
  ggplot(aes(Plate, value)) +
  geom_boxplot() +
  geom_point(size = 2, alpha = 0.8, aes(color = Treatment)) +
  facet_wrap(~ PC, scales = "free")
vf.cell.neg.covar %>% 
  select(PC1:PC5, S1:S3) %>% 
  ggcorr(label = TRUE)
vf.cell.neg.covar %>% 
  select(PC1:PC5, S1:S3) %>% 
  ggpairs()
vf.cell.neg.covar %>% 
  ggplot(aes(PC2, S1)) +
  geom_point(size = 2, alpha = 0.8)
vf.cell.neg.covar %>% 
  select(PC1:PC5, prot_conc_ug_uL:run_order) %>% 
  ggcorr(label = TRUE)
vf.cell.neg.covar %>% 
  select(PC1:PC5, prot_conc_ug_uL:run_order) %>% 
  ggpairs()
vf.cell.neg.covar %>% 
  ggplot(aes(run_order, PC2, color = Plate)) +
  geom_point(size = 4, alpha = 0.8)
vf.cell.neg.covar %>% 
  ggplot(aes(run_order, PC2, color = Treatment)) +
  geom_point(size = 4, alpha = 0.8)
vf.cell.neg.covar %>% 
  ggplot(aes(Treatment, prot_conc_ug_uL, color = Treatment)) +
  geom_boxplot() +
  geom_point(size = 2, alpha = 0.8)

summary(aov(prot_conc_ug_uL ~ Treatment, data = vf.cell.neg.covar))

vf.cell.neg.covar %>% 
  ggplot(aes(prot_conc_ug_uL, PC1, color = Treatment)) +
  geom_point(size = 3, alpha = 0.8)
vf.cell.neg.covar %>% 
  select(S1:prot_conc_ug_uL) %>% 
  ggcorr(label = TRUE)

vf.cell.neg.covar %>% 
  select(S1:prot_conc_ug_uL) %>% 
  ggpairs()
vf.cell.neg.covar %>% 
  ggplot(aes(run_order, S1, color = Treatment)) +
  geom_point(size = 3, alpha = 0.8)
vf.cell.neg.covar %>% 
  ggplot(aes(run_order, S3, color = Treatment)) +
  geom_point(size = 3, alpha = 0.8)
vf.cell.neg.covar %>% 
  ggplot(aes(prot_conc_ug_uL, S2, color = Treatment)) +
  geom_point(size = 3, alpha = 0.8)
vf.cell.neg.covar %>% 
  ggplot(aes(prot_conc_ug_uL, S3, color = Treatment)) +
  geom_point(size = 3, alpha = 0.8)

colnames(vf.cell.neg.mod.vf) <- c("cntrl", "fa_only", "vpa_only", "fa_and_vpa")
# combine the full model matrix and the surrogate variables into one
vf.cell.neg.design.sv <- cbind(vf.cell.neg.mod.vf, vf.cell.neg.surr.var[, 1:2])
vf.cell.neg.cont.mat <- makeContrasts(
  vpa_lowFA = vpa_only - cntrl, 
  vpa_highFA = fa_and_vpa - fa_only,
  FA_diff = (fa_and_vpa - fa_only) - (vpa_only - cntrl),
  levels = c("cntrl", "fa_only", "vpa_only", "fa_and_vpa", "S1", "S2")
  )


vf.cell.neg.fa.cont.mat <- makeContrasts(
  lowFA_cntrl = fa_only - cntrl, 
  highFA_cntrl = fa_and_vpa - vpa_only,
  FA_diff = (fa_and_vpa - vpa_only) - (fa_only - cntrl),
  levels = c("cntrl", "fa_only", "vpa_only", "fa_and_vpa", "S1", "S2")
  )



# fit the model/design matrix
vf.cell.neg.eb <- vf.cell.neg.data.for.sva %>% 
  lmFit(vf.cell.neg.design.sv) %>% 
  contrasts.fit(vf.cell.neg.cont.mat) %>% 
  eBayes()
# pull out the results for each metabolite for each comparison

vf.cell.fa.neg.eb <- vf.cell.neg.data.for.sva %>% 
  lmFit(vf.cell.neg.design.sv) %>% 
  contrasts.fit(vf.cell.neg.fa.cont.mat) %>% 
  eBayes()


vf.cell.neg.eb.tidy <- tidy(vf.cell.neg.eb) %>% 
  mutate(
    adj_pval = p.adjust(p.value, method = "bonferroni"),
    FC = 2 ^ estimate,
    change_in_vpa = ifelse(FC < 1, "down", "up")
    ) %>% 
  rename(compound_short = gene)

vf.cell.fa.neg.eb.tidy <- tidy(vf.cell.fa.neg.eb) %>% 
  mutate(
    adj_pval = p.adjust(p.value, method = "bonferroni"),
    FC = 2 ^ estimate,
    change_in_fa = ifelse(FC < 1, "down", "up")
    ) %>% 
  rename(compound_short = gene)

vf.cell.fa.neg.eb.tidy %>% filter(adj_pval < 0.05)%>% 
  filter(adj_pval < 0.05 & (FC > 1.2 | FC < 1/1.2)) %>% 
  inner_join(vf.cell.neg.compound.info, by = "compound_short")

# volcano plot
vf.cell.neg.eb.tidy %>% 
  ggplot(aes(estimate, -log10(adj_pval))) +
  geom_point(size = 2, alpha = 0.5) +
  geom_hline(linetype = "dashed", color = "#009E73", yintercept = -log10(0.05)) +
  geom_vline(linetype = "dashed", color = "#CC79A7", xintercept = log2(1.2)) +
  geom_vline(linetype = "dashed", color = "#CC79A7", xintercept = log2(1/1.2)) +
  xlim(-2, 2)  +
  xlab("log2(FC)") +
  ylab("-log10(adjusted p-value)") +
  ggtitle("Volcano plot\nVPA + FA / Cells / Negative Mode")
# select statistically significant hits with a certain FC:
vf.cell.neg.hits <- vf.cell.neg.eb.tidy %>% 
  filter(adj_pval < 0.05 & (FC > 1.2 | FC < 1/1.2)) %>% 
  inner_join(vf.cell.neg.compound.info, by = "compound_short")
# how many metabolites are significant across the different contrats
table(vf.cell.neg.hits$term)
# significant metabolites
sort(unique(vf.cell.neg.hits$compound_full))

vf.cell.neg.hits.tally2 <- vf.cell.neg.hits %>% 
  group_by(compound_short, compound_full) %>% 
  count() %>% 
  filter(n == 2)
vf.cell.neg.lowFA.hits <- vf.cell.neg.hits %>% 
  filter(term == "vpa_lowFA" & !(compound_short %in% vf.cell.neg.hits.tally2$compound_short))
vf.cell.neg.highFA.hits <- vf.cell.neg.hits %>% 
  filter(term == "vpa_highFA" & !(compound_short %in% vf.cell.neg.hits.tally2$compound_short))
vf.cell.neg.both.hits <- vf.cell.neg.hits %>% 
  filter(compound_short %in% vf.cell.neg.hits.tally2$compound_short) %>% 
  arrange(compound_short, term)
vf.cell.neg.both.hits %>% 
  select(compound_full, term, FC) %>% 
  spread(key = "term", value = "FC") %>% 
  ggplot(aes(vpa_lowFA, vpa_highFA)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_abline(intercept = 0, slope = 1, color = "blue", alpha = 0.8)
vf.cell.neg.both.hits %>% 
  select(compound_full, term, FC) %>% 
  spread(key = "term", value = "FC") %>% 
  mutate(diff = vpa_highFA - vpa_lowFA) %>% 
  arrange(diff)

#write_csv(path = "./results/vf_cell_neg_all_hits.csv", vf.cell.neg.hits)
```

```{r fig.width=8, fig.height=10, comment=NA}
### Plotting ###
vf.cell.neg.gathered <- vf.cell.neg.noNA %>%
  filter(Group == "sample") %>% 
  bind_cols(vf.cell.neg.surr.var) %>% 
  select(Samples, VPA, FA, S1:S2, starts_with("VPA_FAnC")) %>% 
  gather(key = "Compound", value = "Abundance", VPA_FAnC10:VPA_FAnC99)
vf.cell.neg.nested <- vf.cell.neg.gathered %>% 
  group_by(Compound) %>% 
  nest() %>% 
  mutate(model = map(data, ~lm(Abundance ~ S1 + S2, data = .))) %>% 
  mutate(augment_model = map(model, augment))
vf.cell.neg.modSV.resid <- vf.cell.neg.nested %>% 
  unnest(data, augment_model) %>% 
  select(Samples, VPA, FA, Compound, .resid) %>% 
  spread(Compound, .resid)
vf.cell.neg.modSV.resid %>% 
  select(Samples:FA, one_of(unique(vf.cell.neg.hits$compound_short))) %>% 
  HeatmapPrepAlt("VPA_FAnC") %>% 
  t() %>% 
  heatmaply(
    colors = viridis(n = 10, option = "magma"), 
    xlab = "Samples", ylab = "Compounds",
    main = "Statistically significant compounds\nVPA + FA Experiment / Cells / Neg Mode",
    margins = c(50, 50, 75, 30),
    k_col = 2, k_row = 2,
    labRow = unique(vf.cell.neg.hits$compound_full)
    )
```

```{r resid_pca_cell_neg, comment=NA}
### PCA - cleaned data ###
vf.cell.neg.modSV.pca <- vf.cell.neg.modSV.resid %>% 
  select(starts_with("VPA_FAnC")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
vf.cell.neg.modSV.pca.x <- as.data.frame(vf.cell.neg.modSV.pca$x)
row.names(vf.cell.neg.modSV.pca.x) <- vf.cell.neg.modSV.resid$Samples
vf.cell.neg.modSV.pca.x <- vf.cell.neg.modSV.pca.x %>% 
  bind_cols(vf.cell.neg.modSV.resid %>% select(VPA:FA))
vf.cell.neg.modSV.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, color = VPA, shape = FA)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (56.3% Var)") +
  ylab("PC2 (13.3% Var)")
vf.cell.neg.modSV.pca.x %>% 
  ggplot(aes(x = PC3, y = PC4, color = VPA, shape = FA)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC3 (10.8% Var)") +
  ylab("PC4 (4.7% Var)")
```


## Positive Mode

```{r vpa_fa_cells_pos_stat, comment=NA}
# sample prep
vf.cell.pos.smpl.data <- vf.cell.pos.noNA %>% 
  filter(Group == "sample") %>% 
  unite("Treatment", VPA:FA, sep = "_")
vf.cell.pos.data.for.sva <- as.matrix(
  vf.cell.pos.smpl.data[, which(
    colnames(vf.cell.pos.smpl.data) == "VPA_FApC10"
    ):ncol(vf.cell.pos.smpl.data)]
  )
row.names(vf.cell.pos.data.for.sva) <- vf.cell.pos.smpl.data$Samples
vf.cell.pos.data.for.sva <- t(vf.cell.pos.data.for.sva)
# pheno prep
vf.cell.pos.data.pheno <- as.data.frame(vf.cell.pos.smpl.data[, 5:6])
row.names(vf.cell.pos.data.pheno) <- vf.cell.pos.smpl.data$Samples
# sva calculation
vf.cell.pos.mod.vf <- model.matrix(~ 0 + as.factor(Treatment), data = vf.cell.pos.data.pheno)
vf.cell.pos.mod0 <- model.matrix(~ 1, data = vf.cell.pos.data.pheno)
set.seed(2018)
num.sv(vf.cell.pos.data.for.sva, vf.cell.pos.mod.vf, method = "be")
set.seed(2018)
num.sv(vf.cell.pos.data.for.sva, vf.cell.pos.mod.vf, method = "leek")
set.seed(2018)
vf.cell.pos.sv <- sva(vf.cell.pos.data.for.sva, vf.cell.pos.mod.vf, vf.cell.pos.mod0)
# extract the surrogate variables
vf.cell.pos.surr.var <- as.data.frame(vf.cell.pos.sv$sv)
colnames(vf.cell.pos.surr.var) <- c("S1", "S2")

vf.cell.pos.covar <- vf.cell.pos.smpl.pca.x %>% 
  select(Samples, PC1:PC5) %>% 
  inner_join(
    cbind(vf.cell.pos.data.pheno, vf.cell.pos.surr.var) %>% 
      rownames_to_column("Samples"),
    by = "Samples"
    ) %>% 
  full_join(vf.cell.other.info, by = "Samples") %>% 
  as_tibble()
vf.cell.pos.covar %>% 
  select(Samples, Plate:S2) %>% 
  gather("surr_var", "value", S1:S2) %>% 
  ggplot(aes(Plate, value, color = Plate)) +
  geom_boxplot() +
  geom_jitter(size = 2, alpha = 0.8, width = 0.1) +
  facet_wrap(~ surr_var)
vf.cell.pos.covar %>% 
  select(Samples, Treatment:S2) %>% 
  gather("surr_var", "value", S1:S2) %>% 
  ggplot(aes(Plate, value)) +
  geom_point(size = 3, alpha = 0.8, aes(color = Treatment)) +
  facet_wrap(~ surr_var)
vf.cell.pos.covar %>% 
  select(Samples:Plate) %>% 
  gather("PC", "value", PC1:PC5) %>% 
  ggplot(aes(Plate, value)) +
  geom_boxplot() +
  geom_point(size = 2, alpha = 0.8, aes(color = Treatment)) +
  facet_wrap(~ PC, scales = "free")
vf.cell.pos.covar %>% 
  select(PC1:PC5, S1:S2) %>% 
  ggcorr(label = TRUE)
vf.cell.pos.covar %>% 
  select(PC1:PC5, S1:S2) %>% 
  ggpairs()
vf.cell.pos.covar %>% 
  ggplot(aes(PC2, S1)) +
  geom_point(size = 2, alpha = 0.8)
vf.cell.pos.covar %>% 
  select(PC1:PC5, run_order:prot_conc_ug_uL) %>% 
  ggcorr(label = TRUE)
vf.cell.pos.covar %>% 
  select(PC1:PC5, run_order:prot_conc_ug_uL) %>% 
  ggpairs()
vf.cell.pos.covar %>% 
  ggplot(aes(run_order, PC2, color = Plate)) +
  geom_point(size = 4, alpha = 0.8)
vf.cell.pos.covar %>% 
  ggplot(aes(run_order, PC2, color = Treatment)) +
  geom_point(size = 4, alpha = 0.8)
vf.cell.pos.covar %>% 
  ggplot(aes(prot_conc_ug_uL, PC2, color = Treatment)) +
  geom_point(size = 3, alpha = 0.8)
vf.cell.pos.covar %>% 
  select(S1:prot_conc_ug_uL) %>% 
  ggcorr(label = TRUE)
vf.cell.pos.covar %>% 
  select(S1:prot_conc_ug_uL) %>% 
  ggpairs()
vf.cell.pos.covar %>% 
  ggplot(aes(run_order, S1, color = Treatment)) +
  geom_point(size = 3, alpha = 0.8)

colnames(vf.cell.pos.mod.vf) <- c("cntrl", "fa_only", "vpa_only", "fa_and_vpa")
# combine the full model matrix and the surrogate variables into one
vf.cell.pos.design.sv <- cbind(vf.cell.pos.mod.vf, vf.cell.pos.surr.var)
vf.cell.pos.cont.mat <- makeContrasts(
  vpa_lowFA = vpa_only - cntrl, 
  vpa_highFA = fa_and_vpa - fa_only,
  FA_diff = (fa_and_vpa - fa_only) - (vpa_only - cntrl),
  levels = c("cntrl", "fa_only", "vpa_only", "fa_and_vpa", "S1", "S2")
  )
# fit the model/design matrix
vf.cell.pos.eb <- vf.cell.pos.data.for.sva %>% 
  lmFit(vf.cell.pos.design.sv) %>% 
  contrasts.fit(vf.cell.pos.cont.mat) %>% 
  eBayes()
# pull out the results for each metabolite for each comparison
vf.cell.pos.eb.tidy <- tidy(vf.cell.pos.eb) %>% 
  mutate(
    adj_pval = p.adjust(p.value, method = "bonferroni"),
    FC = 2 ^ estimate,
    change_in_vpa = ifelse(FC < 1, "down", "up")
    ) %>% 
  rename(compound_short = gene)
# volcano plot
vf.cell.pos.eb.tidy %>% 
  ggplot(aes(estimate, -log10(adj_pval))) +
  geom_point(size = 2, alpha = 0.5) +
  geom_hline(linetype = "dashed", color = "#009E73", yintercept = -log10(0.05)) +
  geom_vline(linetype = "dashed", color = "#CC79A7", xintercept = log2(1.2)) +
  geom_vline(linetype = "dashed", color = "#CC79A7", xintercept = log2(1/1.2)) +
  xlim(-2.5, 2.5) +
  xlab("log2(FC)") +
  ylab("-log10(adjusted p-value)") +
  ggtitle("Volcano plot\nVPA + FA / Cells / Positive Mode")
# select statistically significant hits with a certain FC:
vf.cell.pos.hits <- vf.cell.pos.eb.tidy %>% 
  filter(adj_pval < 0.05 & (FC > 1.2 | FC < 1/1.2)) %>% 
  inner_join(vf.cell.pos.compound.info, by = "compound_short")
# how many metabolites are significant across the different contrats
table(vf.cell.pos.hits$term)
# significant metabolites
sort(unique(vf.cell.pos.hits$compound_full))

vf.cell.pos.hits.tally2 <- vf.cell.pos.hits %>% 
  group_by(compound_short, compound_full) %>% 
  count() %>% 
  filter(n == 2)
vf.cell.pos.lowFA.hits <- vf.cell.pos.hits %>% 
  filter(term == "vpa_lowFA" & !(compound_short %in% vf.cell.pos.hits.tally2$compound_short))
vf.cell.pos.highFA.hits <- vf.cell.pos.hits %>% 
  filter(term == "vpa_highFA" & !(compound_short %in% vf.cell.pos.hits.tally2$compound_short))
vf.cell.pos.both.hits <- vf.cell.pos.hits %>% 
  filter(compound_short %in% vf.cell.pos.hits.tally2$compound_short) %>% 
  arrange(compound_short, term)
vf.cell.pos.both.hits %>% 
  select(compound_full, term, FC) %>% 
  spread(key = "term", value = "FC") %>% 
  ggplot(aes(vpa_lowFA, vpa_highFA)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_abline(intercept = 0, slope = 1, color = "blue", alpha = 0.8)
vf.cell.pos.both.hits %>% 
  select(compound_full, term, FC) %>% 
  spread(key = "term", value = "FC") %>% 
  mutate(diff = vpa_highFA - vpa_lowFA) %>% 
  arrange(desc(diff))
#write_csv(path = "./results/vf_cell_pos_all_hits.csv", vf.cell.pos.hits)
```

```{r fig.width=8, fig.height=10, comment=NA}
### Plotting ###
vf.cell.pos.gathered <- vf.cell.pos.noNA %>%
  filter(Group == "sample") %>% 
  bind_cols(vf.cell.pos.surr.var) %>% 
  select(Samples, VPA, FA, S1:S2, starts_with("VPA_FApC")) %>% 
  gather(key = "Compound", value = "Abundance", VPA_FApC10:VPA_FApC99)
vf.cell.pos.nested <- vf.cell.pos.gathered %>% 
  group_by(Compound) %>% 
  nest() %>% 
  mutate(model = map(data, ~lm(Abundance ~ S1 + S2, data = .))) %>% 
  mutate(augment_model = map(model, augment))
vf.cell.pos.modSV.resid <- vf.cell.pos.nested %>% 
  unnest(data, augment_model) %>% 
  select(Samples, VPA, FA, Compound, .resid) %>% 
  spread(Compound, .resid)
vf.cell.pos.modSV.resid %>% 
  select(Samples:FA, one_of(unique(vf.cell.pos.hits$compound_short))) %>% 
  HeatmapPrepAlt("VPA_FApC") %>% 
  t() %>% 
  heatmaply(
    colors = viridis(n = 10, option = "magma"), 
    xlab = "Samples", ylab = "Compounds",
    main = "Statistically significant compounds\nVPA + FA Experiment / Cells / Positive Mode",
    margins = c(50, 50, 75, 30),
    k_col = 2, k_row = 2,
    labRow = unique(vf.cell.pos.hits$compound_full)
    )
```

```{r resid_pca_cell_pos, comment=NA}
### PCA - cleaned data ###
vf.cell.pos.modSV.pca <- vf.cell.pos.modSV.resid %>% 
  select(starts_with("VPA_FApC")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
vf.cell.pos.modSV.pca.x <- as.data.frame(vf.cell.pos.modSV.pca$x)
row.names(vf.cell.pos.modSV.pca.x) <- vf.cell.pos.modSV.resid$Samples
vf.cell.pos.modSV.pca.x <- vf.cell.pos.modSV.pca.x %>% 
  bind_cols(vf.cell.pos.modSV.resid %>% select(VPA:FA))
vf.cell.pos.modSV.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, color = VPA, shape = FA)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (60.6% Var)") +
  ylab("PC2 (11.6% Var)")
vf.cell.pos.modSV.pca.x %>% 
  ggplot(aes(x = PC3, y = PC4, color = VPA, shape = FA)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC3 (7.4% Var)") +
  ylab("PC4 (4.5% Var)")
```


## Write to csv

(not evaluated)

```{r modsv_to_csv, comment=NA, eval = FALSE}
#write_csv(vf.cell.neg.modSV.resid, path = "./results/modSV_resid/vpa_fa_exp1_cell_neg_modSV_resid.csv")
#write_csv(vf.cell.pos.modSV.resid, path = "./results/modSV_resid/vpa_fa_exp1_cell_pos_modSV_resid.csv")
#write_csv(vf.med.neg.modSV.resid, path = "./results/modSV_resid/vpa_fa_exp1_media_neg_modSV_resid.csv")
#write_csv(vf.med.pos.modSV.resid, path = "./results/modSV_resid/vpa_fa_exp1_media_pos_modSV_resid.csv")
```


# Profile Plot

```{r comment=NA, eval = FALSE}
vf.cell.neg.hits <- vf.cell.neg.hits %>% 
  filter(compound_short != "VPA_FAnC49")

vf.cell.neg.resid.for.profile <- vf.cell.neg.modSV.resid %>% 
  select(Samples:FA, one_of(vf.cell.neg.hits$compound_short)) %>% 
  unite("Treatment", VPA:FA, sep = "_")

vf.cell.neg.resid.for.profile %>% 
  gather("Compound", value = "Abundance", -c(Samples, Treatment)) %>% 
  group_by(Samples) %>% 
  count() %>% 
  filter(n != 36)

vf.cell.neg.resid.order <- vf.cell.neg.resid.for.profile %>% 
  gather("Compound", value = "Abundance", -c(Samples, Treatment)) %>% 
  group_by(Compound, Treatment) %>% 
  summarize(avg.abun = mean(Abundance)) %>% 
  ungroup() %>% 
  spread(key = "Treatment", value = "avg.abun") %>% 
  arrange(cntrl_cntrl) %>% 
  mutate(compound_sort = factor(Compound)) %>% 
  inner_join(
    vf.cell.neg.hits %>% 
      select(compound_short, compound_full:cas_id) %>% 
      distinct(), 
    by = c("Compound" = "compound_short")
    )
  

vf.cell.neg.modSV.resid.prof.plot <- vf.cell.neg.resid.for.profile %>% 
  gather("Compound", value = "Abundance", -c(Samples, Treatment)) %>% 
  mutate(compound_order = factor(Compound, levels = vf.cell.neg.resid.order$compound_sort)) %>% 
  ggplot(aes(compound_order, Abundance, color = Treatment, group = Samples)) + 
  geom_line(alpha = 0.25, size = 1.5) + 
  theme_minimal() +
  theme(
    axis.ticks.x = element_blank(),
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
    ) +
  scale_x_discrete(name = "Compound", labels = vf.cell.neg.resid.order$compound_full) +
  scale_color_manual(values = c("#56B4E9", "#0072B2", "#E69F00",  "#D55E00"), labels = c("Control", "5X FA", "VPA", "VPA + 5X FA")) +
  ylab("log2(Compound Resid)") +
  # overlay the group averages 
  geom_line(
    data = vf.cell.neg.resid.order %>% gather("Treatment", "treat_mean", cntrl_cntrl:vpa_fa), aes(compound_sort, treat_mean, color = Treatment, group = Treatment),
    size = 2
    )
vf.cell.neg.modSV.resid.prof.plot



vf.cell.pos.resid.for.profile <- vf.cell.pos.modSV.resid %>% 
  select(Samples:FA, one_of(vf.cell.pos.hits$compound_short)) %>% 
  unite("Treatment", VPA:FA, sep = "_")

vf.cell.pos.resid.for.profile %>% 
  gather("Compound", value = "Abundance", -c(Samples, Treatment)) %>% 
  group_by(Samples) %>% 
  count() %>% 
  filter(n != 32)

vf.cell.pos.resid.order <- vf.cell.pos.resid.for.profile %>% 
  gather("Compound", value = "Abundance", -c(Samples, Treatment)) %>% 
  group_by(Compound, Treatment) %>% 
  summarize(avg.abun = mean(Abundance)) %>% 
  ungroup() %>% 
  spread(key = "Treatment", value = "avg.abun") %>% 
  arrange(cntrl_cntrl) %>% 
  mutate(compound_sort = factor(Compound)) %>% 
  inner_join(
    vf.cell.pos.hits %>% 
      select(compound_short, compound_full:cas_id) %>% 
      distinct(), 
    by = c("Compound" = "compound_short")
    )
  

vf.cell.pos.modSV.resid.prof.plot <- vf.cell.pos.resid.for.profile %>% 
  gather("Compound", value = "Abundance", -c(Samples, Treatment)) %>% 
  mutate(compound_order = factor(Compound, levels = vf.cell.pos.resid.order$compound_sort)) %>% 
  ggplot(aes(compound_order, Abundance, color = Treatment, group = Samples)) + 
  geom_line(alpha = 0.25, size = 1.5) + 
  theme_minimal() +
  theme(
    axis.ticks.x = element_blank(),
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
    ) +
  scale_x_discrete(name = "Compound", labels = vf.cell.pos.resid.order$compound_full) +
  scale_color_manual(values = c("#56B4E9", "#0072B2", "#E69F00",  "#D55E00"), labels = c("Control", "5X FA", "VPA", "VPA + 5X FA")) +
  ylab("log2(Compound Resid)") +
  # overlay the group averages 
  geom_line(
    data = vf.cell.pos.resid.order %>% gather("Treatment", "treat_mean", cntrl_cntrl:vpa_fa), aes(compound_sort, treat_mean, color = Treatment, group = Treatment),
    size = 2
    )
vf.cell.pos.modSV.resid.prof.plot

vf.cell.resid.prof.grid.plot <- plot_grid(
  vf.cell.neg.modSV.resid.prof.plot,
  vf.cell.pos.modSV.resid.prof.plot,
  labels = c("A", "B"),
  ncol = 1
)

#save_plot("./results/vf_cell_resid_profile_grid_plot.png", vf.cell.resid.prof.grid.plot, base_height = 10, base_width = 8)
```


# Session Info

```{r session_info, comment=NA}
sessionInfo()
```

