---
title: "Lrp6 KO Exp 2"
author: "Darya Akimova"
date: "January 2019"
output: 
  html_document:
    toc: true
    toc_float:
      smooth_scroll: false
    number_sections: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Setup

## Packages

```{r packages, comment=NA, message = FALSE}
library(tidyverse)
library(cowplot)
library(GGally)
library(heatmaply)
library(sva)
library(limma)
library(biobroom)
library(ggridges)
library(ggthemes)
theme_set(theme_minimal())
```


## Data

### Metabolite Abundances

```{r abundance_import, comment=NA}
ko2.neg.raw <- read_csv("./data/abundances/lrp6ko_exp2_abundances_negmode.csv")
ko2.pos.raw <- read_csv("./data/abundances/lrp6ko_exp2_abundances_posmode.csv")
setequal(ko2.neg.raw$Samples, ko2.pos.raw$Samples)
```


### Compound Information

```{r compound_info_import, comment=NA}
ko2.neg.compound.info <- read_csv("./data/compound_info/lrp6ko_exp2_compound_info_negmode.csv")
ko2.pos.compound.info <- read_csv("./data/compound_info/lrp6ko_exp2_compound_info_posmode.csv")
```


## Functions

```{r functions, comment=NA}
MissingPerSamplePlot <- function(raw.data, start.string) {
  # Counts the number of missing/NA values per sample and
  # percent compounds missing out of total number of compounds per sample
  # Then passes the result into a vertical bar plot, where each 
  # bar represents a single sample and the size of the bar 
  # is the % of compounds missing
  
  counted.na <- raw.data %>%
  select(starts_with(start.string)) %>% 
  mutate(
    count.na = apply(., 1, function(x) sum(is.na(x))),
    percent.na = (count.na / ncol(raw.data %>% select(starts_with(start.string)))) * 100
    ) %>%
  dplyr::select(count.na, percent.na) %>%
  bind_cols(
    raw.data %>% 
      select(Samples, Type)
      ) %>% 
  arrange(percent.na) %>% 
  mutate(f.order = factor(Samples, levels = Samples))
counted.na %>% 
  ggplot(aes(x = f.order, y = percent.na, fill = Type)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 20, color = "gray", size = 1, alpha = 0.8) +
  coord_flip()+
  xlab("Samples") +
  ylab("Percent missing values in sample") +
  theme(axis.text.y = element_text(size = 6)) 
}
MissingPerCompound <- function(raw.data, start.string){
  # Function to count in how many experimental samples each compound is missing
  # Also calculates the % missing:
  # (# NA per compound across all experimental samples) * 100 / (tot num of samples)
  
  raw.data %>% 
  filter(Type == "sample") %>% 
  select(Samples, starts_with(start.string)) %>% 
  gather(key = "Compound", value = "Abundance", -Samples) %>% 
  group_by(Compound) %>% 
  summarise(
    na_count = sum(is.na(Abundance)),
    n_samples = n(),
    percent_na = (na_count * 100 / n_samples)
    ) %>% 
  filter(na_count > 0) %>% 
  arrange(desc(percent_na))
}
ReplaceNAwMinLogTransform <- function(raw.dataframe, start.prefix, type) {
  # Function to replace any NAs in each column with the minimum for that column, 
  # separately for each data.set (patient and dabase samples separately)
  #
  # Returns: returns data frame(? check on this) with only the values bound by the function args
  # 
  smpls <- raw.dataframe %>%
    filter(Type == "sample") %>% 
    dplyr::select(starts_with(start.prefix))
  smpls.min <- lapply(smpls, min, na.rm = TRUE)
  # replace the missing values in the real samples with the minimum of the samples
  # then take the log
  smpls.noNA <- raw.dataframe  %>%
    filter(Type == "sample") %>% 
    dplyr::select(Samples:Status) %>%
    bind_cols(
      smpls %>%
        replace_na(replace = smpls.min) %>% 
        log2()
      ) 
  # replace missing QC values with values within the QC groups
  QC <- raw.dataframe %>%
    filter(Type == "QC") %>% 
    dplyr::select(starts_with(start.prefix))
  QC.min <- lapply(QC, min, na.rm = TRUE)
  # deal with cases where there is no min:
  QC.min <- lapply(QC.min, function(x) ifelse(is.infinite(x),2,x))
  QC.noNA <- raw.dataframe  %>%
    filter(Type == "QC") %>% 
    dplyr::select(Samples:Status) %>%
    bind_cols(
      QC %>%
        replace_na(replace = QC.min) %>% 
        log2()
      )
  # replace the missing values in solv and empty samples with 1 - for PCA analysis
  # then take the log
  other.min <- setNames(
    as.list(
      rep(2, ncol(
        raw.dataframe %>% 
          dplyr::select(starts_with(start.prefix))))
      ),
    colnames(raw.dataframe %>% dplyr::select(starts_with(start.prefix)))
                )
  other.num.log <- raw.dataframe %>%
    filter(Type == "Solv") %>% 
    dplyr::select(Samples:Status) %>%
    bind_cols(
      raw.dataframe %>% 
        filter(Type == "Solv") %>% 
        dplyr::select(starts_with(start.prefix)) %>%
        replace_na(replace = other.min) %>% 
        log2()
      )
  # combine them together back into one data frame
  all.noNA <- smpls.noNA %>% 
    bind_rows(QC.noNA) %>% 
    bind_rows(other.num.log)
}
HeatmapPrepAlt <- function(raw.data, start.prefix){
  # function for preparing dara for heatmap viz
  x <- raw.data %>% 
    select(starts_with(start.prefix)) %>% 
    scale(center = TRUE, scale = TRUE) %>% 
    as.matrix() 
  row.names(x) <- raw.data$Samples
  return(x)
}
```


# Data Exploration

## Mass vs Retention Time

Q: What are the distributions of compound masses and retention times?

```{r mass_v_RT_plots, fig.height=7, fig.width=9, comment=NA}
full.ko2.cmpnd <- ko2.neg.compound.info %>% 
  mutate(Set = "Neg") %>% 
  bind_rows(ko2.pos.compound.info %>% mutate(Set = "Pos"))
full.ko2.cmpnd %>% 
  ggplot(aes(x = RT, y = Mass)) +
  geom_point(size = 3, alpha = 0.3) +
  xlab("Retention Time (min)") +
  ylab("Mass (Da)") +
  ggtitle("Mass v RT\nVPA + FA Exp") +
  ylim(0, 1000)
full.ko2.cmpnd %>% 
  ggplot(aes(x = RT, y = Mass, color = Set)) +
  geom_point(size = 3, alpha = 0.8) +
  xlab("Retnetion Time (min)") +
  ylab("Mass (Da)") +
  ggtitle("Mass v RT\nVPA + FA Exp") +
  facet_wrap(~ Set) +
  ylim(0, 1000)
```


Q: Which compounds were found in one or more of the data types?


```{r cmnpd_mode_match, fig.height= 5, fig.width=8, comment=NA}
ko2.cmpnd.join <- ko2.neg.compound.info %>% 
  inner_join(ko2.pos.compound.info, by = "CAS_ID", suffix = c("_n", "_p")) %>% 
  select(
    CAS_ID, contains("short"), 
    contains("full"), starts_with("Formula"), 
    starts_with("Mass"), starts_with("RT")
    )
# compound names - found in pos and neg mode / cells 
print(ko2.cmpnd.join$Compound_full_n)
# percent of cell / neg compounds found in cell / pos 
round(nrow(ko2.cmpnd.join) * 100 / nrow(ko2.neg.compound.info), 1)
# percent of cell / neg compounds found in cell / pos 
round(nrow(ko2.cmpnd.join) * 100 / nrow(ko2.pos.compound.info), 1)
# problem checks #
ko2.cmpnd.join %>% 
  select(contains("Mass")) %>% 
  ggpairs()
ko2.cmpnd.join %>% 
  select(starts_with("RT")) %>% 
  ggpairs()
```


## Missing Values

Q: Do any of the samples have greater than 20% missing (NA) compound abundances, out of all of the features in the dataset?

```{r missing_per_sample_plots, comment=NA}
MissingPerSamplePlot(ko2.neg.raw, "Lrp6KO2n") +
  ggtitle("Missing Per Sample\nLrp6KO / Exp2 / Neg Mode")
MissingPerSamplePlot(ko2.pos.raw, "Lrp6KO2p") +
  ggtitle("Missing Per Sample\nLrp6KO / Exp2 / Pos Mode")
```

I think I had mixed up the QC and Solvent samples in this experiment when running. 

Q: Are any of the compounds more than 20% missing in the experimental sample group? If there are any, they will be excluded from analysis. 

```{r percent_missing_per_compound, comment=NA}
(ko2.neg.cmpnd.to.excl <- MissingPerCompound(ko2.neg.raw, "Lrp6KO2n") %>% 
  filter(percent_na > 20))
(ko2.pos.cmpnd.to.excl <- MissingPerCompound(ko2.pos.raw, "Lrp6KO2p") %>% 
  filter(percent_na > 20))
```


## Negative Controls and Compound Elimination

### Negative Mode

```{r empty_solv_neg, comment=NA}
ko2.neg.raw.type.mean <- ko2.neg.raw %>% 
  group_by(Type) %>% 
  summarize_at(vars(matches("Lrp6KO2n")), mean, na.rm = TRUE) %>% 
  gather(key = "Compound", value = "Type_mean_abun", -Type)
ko2.neg.raw.type.mean %>% 
  ggplot(aes(log2(Type_mean_abun), color = Type)) +
  geom_density(size = 2, alpha = 0.8) +
  ggtitle("Distribution of compound means\nLrp6KO Exp 2 / Negative Mode\nGrouped by sample type")
ko2.neg.raw.type.mean.order <- ko2.neg.raw.type.mean %>% 
  filter(Type == "sample") %>% 
  arrange(Type_mean_abun)
ko2.neg.raw %>% 
  select(Samples, Type, starts_with("Lrp6KO2n")) %>% 
  gather("Compound", value = "Abundance", -c(Samples, Type)) %>% 
  mutate(Cmpnd_sort = factor(Compound, levels = ko2.neg.raw.type.mean.order$Compound)) %>% 
  ggplot(aes(Cmpnd_sort, log2(Abundance), color = Type, group = Samples)) + 
  geom_line(alpha = 0.1, size = 1) + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab("Compound") +
  geom_line(
    data = ko2.neg.raw.type.mean %>% 
      mutate(Cmpnd_sort = factor(Compound, levels = ko2.neg.raw.type.mean.order$Compound)), 
    aes(Cmpnd_sort, log2(Type_mean_abun), color = Type, group = Type),
    size = 0.5
    ) +
  ggtitle("Profile Plot of all compound abundances\nWith average per sample type overlaid\nLrp6KO Exp 2 / Negative Mode")
ko2.neg.raw.type.mean %>% 
  mutate(Cmpnd_sort = factor(Compound, levels = ko2.neg.raw.type.mean.order$Compound)) %>% 
  ggplot(aes(Cmpnd_sort, log2(Type_mean_abun), color = Type, group = Type)) +
  geom_point(size = 1, alpha = 0.8) +
  geom_line(alpha = 0.8) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab("Compound") +
  ylab("log2(Sample Type Mean)") +
  ggtitle("Profile Plot of compound means by sample type only\nLrp6KO Exp 2 / Negative Mode")
ko2.neg.raw.type.diff <- ko2.neg.raw.type.mean %>% 
  spread(Type, Type_mean_abun) %>% 
  mutate(smpl_solv_diff = sample / Solv)
ko2.neg.raw.type.diff %>% 
  ggplot(aes(log2(smpl_solv_diff))) +
  geom_histogram(bins = 50)
# include compounds with FC > 2.5 or FC is NA (indication of NA in solv or empty)
ko2.neg.cmpnd.to.incl <- ko2.neg.raw.type.diff  %>% 
  filter(smpl_solv_diff > 2.5 | is.na(smpl_solv_diff)) %>% 
  filter(!(Compound %in% ko2.neg.cmpnd.to.excl$Compound))
# original number of metabolites
nrow(ko2.neg.raw.type.diff)
# number of metabolites after filtering 
nrow(ko2.neg.cmpnd.to.incl)
```


### Positive Mode

```{r empty_solv_pos, comment=NA}
ko2.pos.raw.type.mean <- ko2.pos.raw %>% 
  group_by(Type) %>% 
  summarize_at(vars(matches("Lrp6KO2p")), mean, na.rm = TRUE) %>% 
  gather(key = "Compound", value = "Type_mean_abun", -Type)
ko2.pos.raw.type.mean %>% 
  ggplot(aes(log2(Type_mean_abun), color = Type)) +
  geom_density(size = 2, alpha = 0.8) +
  ggtitle("Distribution of compound means\nLrp6KO Exp 2 / Positive Mode\nGrouped by sample type")
ko2.pos.raw.type.mean.order <- ko2.pos.raw.type.mean %>% 
  filter(Type == "sample") %>% 
  arrange(Type_mean_abun)
ko2.pos.raw %>% 
  select(Samples, Type, starts_with("Lrp6KO2p")) %>% 
  gather("Compound", value = "Abundance", -c(Samples, Type)) %>% 
  mutate(Cmpnd_sort = factor(Compound, levels = ko2.pos.raw.type.mean.order$Compound)) %>% 
  ggplot(aes(Cmpnd_sort, log2(Abundance), color = Type, group = Samples)) + 
  geom_line(alpha = 0.1, size = 1) + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab("Compound") +
  geom_line(
    data = ko2.pos.raw.type.mean %>% 
      mutate(Cmpnd_sort = factor(Compound, levels = ko2.pos.raw.type.mean.order$Compound)), 
    aes(Cmpnd_sort, log2(Type_mean_abun), color = Type, group = Type),
    size = 0.5
    ) +
  ggtitle("Profile Plot of all compound abundances\nWith average per sample type overlaid\nLrp6KO Exp 2 / Positive Mode")
ko2.pos.raw.type.mean %>% 
  mutate(Cmpnd_sort = factor(Compound, levels = ko2.pos.raw.type.mean.order$Compound)) %>% 
  ggplot(aes(Cmpnd_sort, log2(Type_mean_abun), color = Type, group = Type)) +
  geom_point(size = 1, alpha = 0.8) +
  geom_line(alpha = 0.8) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  xlab("Compound") +
  ylab("log2(Sample Type Mean)") +
  ggtitle("Profile Plot of compound means by sample type only\nLrp6KO Exp 2 / Positive Mode")
ko2.pos.raw.type.diff <- ko2.pos.raw.type.mean %>% 
  spread(Type, Type_mean_abun) %>% 
  mutate(smpl_solv_diff = sample / Solv)
ko2.pos.raw.type.diff %>% 
  ggplot(aes(log2(smpl_solv_diff))) +
  geom_histogram(bins = 50)
# include compounds with FC > 2.5 or FC is NA (indication of NA in solv or empty)
ko2.pos.cmpnd.to.incl <- ko2.pos.raw.type.diff  %>% 
  filter(smpl_solv_diff > 2.5 | is.na(smpl_solv_diff)) %>% 
  filter(!(Compound %in% ko2.pos.cmpnd.to.excl$Compound))
# original number of metabolites
nrow(ko2.pos.raw.type.diff)
# number of metabolites after filtering 
nrow(ko2.pos.cmpnd.to.incl)
```


# Data Prep and Preliminary Analysis

## Cleanup

```{r min_na_rpl_log_transform, comment=NA}
ko2.neg.noNA <- ko2.neg.raw %>% 
  select(Samples:Status, one_of(ko2.neg.cmpnd.to.incl$Compound)) %>% 
  ReplaceNAwMinLogTransform("Lrp6KO2n")
ko2.pos.noNA <- ko2.pos.raw %>% 
  select(Samples:Status, one_of(ko2.pos.cmpnd.to.incl$Compound)) %>% 
  ReplaceNAwMinLogTransform("Lrp6KO2p")
```


## Distribution plots

### Negative Mode

```{r distribution_plots_neg, comment=NA}
ko2.neg.noNA.gathered <- ko2.neg.noNA %>% 
  gather(
    key = "Metabolite", "Abundance", 
    which(colnames(ko2.neg.noNA) == "Lrp6KO2n1"):ncol(ko2.neg.noNA)
    )
# plot all abundances in a sample, grouped by sample as a boxplot
ko2.neg.noNA.gathered %>% 
  ggplot(aes(Samples, Abundance, fill = Type)) +
  geom_boxplot() +
  geom_boxplot(aes(color = Type), fatten = NULL, fill = NA, coef = 0, outlier.alpha = 0, show.legend = FALSE) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("log2(Abundance)") +
  ggtitle("Boxplot of compound abundances\nAll samples\nLrp6KO Exp 2/ Negative Mode")
# same data format, but as ridge plots
ko2.neg.noNA.gathered %>% 
  ggplot(aes(y = Samples, x = Abundance, fill = Type)) + 
  geom_density_ridges(scale = 15) +
  theme_ridges() +
  scale_y_discrete(expand = c(0.01, 0)) +
  ggtitle("Ridge plot of compound abundances\nAll samples\nLrp6KO Exp 2/ Negative Mode")
# experimental samples only
ko2.neg.noNA.gathered %>% 
  filter(Type == "sample") %>% 
  unite("Condition", Folate:Status, sep = "_") %>% 
  ggplot(aes(y = Samples, x = Abundance, fill = Condition)) + 
  geom_density_ridges(scale = 10) +
  theme_ridges() +
  scale_y_discrete(expand = c(0.01, 0)) +
  ggtitle("Ridge plot of compound abundances\nExperimental samples only\nLrp6KO Exp 2/ Negative Mode")
# overlay the distributions for another look
ko2.neg.noNA.gathered %>%
  filter(Type == "sample") %>% 
  unite("Condition", Folate:Status, sep = "_") %>% 
  ggplot(aes(Abundance, group = Samples, color = Condition)) +
  geom_density(alpha = 0.8, size = 0.75) +
  xlab("log2(Abundance)") +
  ggtitle("Density plot of compound abundances\nExperimental samples only\nLrp6KO Exp 2/ Negative Mode")
```


### Positive Mode

```{r distribution_plots_pos, comment=NA}
ko2.pos.noNA.gathered <- ko2.pos.noNA %>% 
  gather(
    key = "Metabolite", "Abundance", 
    which(colnames(ko2.pos.noNA) == "Lrp6KO2p10"):ncol(ko2.pos.noNA)
    )
# plot all abundances in a sample, grouped by sample as a boxplot
ko2.pos.noNA.gathered %>% 
  ggplot(aes(Samples, Abundance, fill = Type)) +
  geom_boxplot() +
  geom_boxplot(aes(color = Type), fatten = NULL, fill = NA, coef = 0, outlier.alpha = 0, show.legend = FALSE) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("log2(Abundance)") +
  ggtitle("Boxplot of compound abundances\nAll samples\nLrp6KO Exp 2/ Positive Mode")
# same data format, but as ridge plots
ko2.pos.noNA.gathered %>% 
  ggplot(aes(y = Samples, x = Abundance, fill = Type)) + 
  geom_density_ridges(scale = 15) +
  theme_ridges() +
  scale_y_discrete(expand = c(0.01, 0)) +
  ggtitle("Ridge plot of compound abundances\nAll samples\nLrp6KO Exp 2/ Positive Mode")
# experimental samples only
ko2.pos.noNA.gathered %>% 
  filter(Type == "sample") %>% 
  unite("Condition", Folate:Status, sep = "_") %>% 
  ggplot(aes(y = Samples, x = Abundance, fill = Condition)) + 
  geom_density_ridges(scale = 10) +
  theme_ridges() +
  scale_y_discrete(expand = c(0.01, 0)) +
  ggtitle("Ridge plot of compound abundances\nExperimental samples only\nLrp6KO Exp 2/ Positive Mode")
# overlay the distributions for another look
ko2.pos.noNA.gathered %>%
  filter(Type == "sample") %>% 
  unite("Condition", Folate:Status, sep = "_") %>% 
  ggplot(aes(Abundance, group = Samples, color = Condition)) +
  geom_density(alpha = 0.8, size = 0.75) +
  xlab("log2(Abundance)") +
  ggtitle("Density plot of compound abundances\nExperimental samples only\nLrp6KO Exp 2/ Positive Mode")
```


## Principal Component Analysis 

Some plots to understand the relationship between the samples, QC samples, solvent, and empty samples in some cases. 

### Negative Mode 

```{r pca_neg, comment=NA}
### PCA on all Samples ###
ko2.neg.full.pca <- ko2.neg.noNA %>% 
  select(starts_with("Lrp6KO2n")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (ko2.neg.full.pca$sdev ^ 2) * 100 / sum(ko2.neg.full.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nAll samples only\nLrp6KO Exp 2 / Negative Mode",
  type = "b"
  )
ko2.neg.full.pca.x <- as.data.frame(ko2.neg.full.pca$x)
row.names(ko2.neg.full.pca.x) <- ko2.neg.noNA$Samples
ko2.neg.full.pca.x <- ko2.neg.full.pca.x %>% 
  bind_cols(ko2.neg.noNA %>% select(Type:Status))
ko2.neg.full.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, color = Type)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (79.0% Var)") +
  ylab("PC2 (10.6% Var)") +
  ggtitle("Principal Component Analysis\nAll Samples\nLrp6KO Exp 2 / Negative Mode") +
  scale_color_tableau()

### Samples and QC ###
ko2.neg.smpl.QC.pca <- ko2.neg.noNA %>% 
  filter(Type == "sample" | Type == "QC") %>% 
  select(starts_with("Lrp6KO2n")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (ko2.neg.smpl.QC.pca$sdev ^ 2) * 100 / sum(ko2.neg.smpl.QC.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nSamples and QC\nLrp6KO Exp 2 / Negative Mode",
  type = "b"
  )
ko2.neg.smpl.QC.pca.x <- as.data.frame(ko2.neg.smpl.QC.pca$x)
ko2.neg.smpl.QC.pca.x <- ko2.neg.smpl.QC.pca.x %>% 
  bind_cols(
    ko2.neg.noNA %>% 
      filter(Type == "sample" | Type == "QC") %>% 
      select(Samples, Type:Status)
  )
row.names(ko2.neg.smpl.QC.pca.x) <- ko2.neg.smpl.QC.pca.x$Samples
ko2.neg.smpl.QC.pca.x %>% 
  unite("Condition", Folate:Status, sep = "_") %>% 
  ggplot(aes(x = PC1, y = PC2, color = Condition)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (79.2% Var)") +
  ylab("PC2 (7.0% Var)") +
  ggtitle("Principal Component Analysis\nSamples and QC\nLrp6KO Exp 2 / Negative Mode") +
  scale_color_tableau()
ko2.neg.smpl.QC.pca.x %>% 
  unite("Condition", Folate:Status, sep = "_") %>% 
  ggplot(aes(x = PC3, y = PC4, color = Condition)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC3 (4.1% Var)") +
  ylab("PC4 (1.7% Var)") +
  ggtitle("Principal Component Analysis\nSamples and QC\nLrp6KO Exp 2 / Negative Mode") +
  scale_color_tableau()

### Experimental Samples Only ###
ko2.neg.smpl.pca <- ko2.neg.noNA %>% 
  filter(Type == "sample") %>% 
  select(starts_with("Lrp6KO2")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (ko2.neg.smpl.pca$sdev ^ 2) * 100 / sum(ko2.neg.smpl.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nExperimental samples only\nLrp6KO Exp 2 / Negative Mode",
  type = "b"
  )
ko2.neg.smpl.pca.x <- as.data.frame(ko2.neg.smpl.pca$x)
ko2.neg.smpl.pca.x <- ko2.neg.smpl.pca.x %>% 
  bind_cols(
    ko2.neg.noNA %>% 
      filter(Type == "sample") %>% 
      select(Samples, Type:Status)
  )
row.names(ko2.neg.smpl.pca.x) <- ko2.neg.smpl.pca.x$Samples
ko2.neg.smpl.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, shape = Folate, color = Status)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (47.1% Var)") +
  ylab("PC2 (10.4% Var)") +
  ggtitle("Principal Component Analysis\nExperimental samples only\nLrp6KO Exp 2 / Negative Mode") +
  scale_color_tableau()
ko2.neg.smpl.pca.x %>% 
  ggplot(aes(x = PC3, y = PC4, shape = Folate, color = Status)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC3 (7.7% Var)") +
  ylab("PC4 (6.7% Var)") +
  ggtitle("Principal Component Analysis\nExperimental samples only\nLrp6KO Exp 2 / Negative Mode") +
  scale_color_tableau()
```


### Positive Mode 

```{r pca_pos, comment=NA}
### PCA on all Samples ###
ko2.pos.full.pca <- ko2.pos.noNA %>% 
  select(starts_with("Lrp6KO2p")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (ko2.pos.full.pca$sdev ^ 2) * 100 / sum(ko2.pos.full.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nAll samples only\nLrp6KO Exp 2 / Positive Mode",
  type = "b"
  )
ko2.pos.full.pca.x <- as.data.frame(ko2.pos.full.pca$x)
row.names(ko2.pos.full.pca.x) <- ko2.pos.noNA$Samples
ko2.pos.full.pca.x <- ko2.pos.full.pca.x %>% 
  bind_cols(ko2.pos.noNA %>% select(Type:Status))
ko2.pos.full.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, color = Type)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (89.9% Var)") +
  ylab("PC2 (1.5% Var)") +
  ggtitle("Principal Component Analysis\nAll Samples\nLrp6KO Exp 2 / Positive Mode") +
  scale_color_tableau()

### Samples and QC ###
ko2.pos.smpl.QC.pca <- ko2.pos.noNA %>% 
  filter(Type == "sample" | Type == "QC") %>% 
  select(starts_with("Lrp6KO2p")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (ko2.pos.smpl.QC.pca$sdev ^ 2) * 100 / sum(ko2.pos.smpl.QC.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nSamples and QC\nLrp6KO Exp 2 / Positive Mode",
  type = "b"
  )
ko2.pos.smpl.QC.pca.x <- as.data.frame(ko2.pos.smpl.QC.pca$x)
ko2.pos.smpl.QC.pca.x <- ko2.pos.smpl.QC.pca.x %>% 
  bind_cols(
    ko2.pos.noNA %>% 
      filter(Type == "sample" | Type == "QC") %>% 
      select(Samples, Type:Status)
  )
row.names(ko2.pos.smpl.QC.pca.x) <- ko2.pos.smpl.QC.pca.x$Samples
ko2.pos.smpl.QC.pca.x %>% 
  unite("Condition", Folate:Status, sep = "_") %>% 
  ggplot(aes(x = PC1, y = PC2, color = Condition)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (35.0% Var)") +
  ylab("PC2 (16.7% Var)") +
  ggtitle("Principal Component Analysis\nSamples and QC\nLrp6KO Exp 2 / Positive Mode") +
  scale_color_tableau()
ko2.pos.smpl.QC.pca.x %>% 
  unite("Condition", Folate:Status, sep = "_") %>% 
  ggplot(aes(x = PC3, y = PC4, color = Condition)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC3 (12.3% Var)") +
  ylab("PC4 (8.0% Var)") +
  ggtitle("Principal Component Analysis\nSamples and QC\nLrp6KO Exp 2 / Positive Mode") +
  scale_color_tableau()

### Experimental Samples Only ###
ko2.pos.smpl.pca <- ko2.pos.noNA %>% 
  filter(Type == "sample") %>% 
  select(starts_with("Lrp6KO2")) %>% 
  mutate_all(scale, center = TRUE, scale = FALSE) %>% 
  as.matrix() %>% 
  prcomp()
plot(
  (ko2.pos.smpl.pca$sdev ^ 2) * 100 / sum(ko2.pos.smpl.pca$sdev ^ 2), 
  xlab = "Principal Component",
  ylab = "Variance Explained",
  main = "Percent variance explained by each principal component\nExperimental samples only\nLrp6KO Exp 2 / Positive Mode",
  type = "b"
  )
ko2.pos.smpl.pca.x <- as.data.frame(ko2.pos.smpl.pca$x)
ko2.pos.smpl.pca.x <- ko2.pos.smpl.pca.x %>% 
  bind_cols(
    ko2.pos.noNA %>% 
      filter(Type == "sample") %>% 
      select(Samples, Type:Status)
  )
row.names(ko2.pos.smpl.pca.x) <- ko2.pos.smpl.pca.x$Samples
ko2.pos.smpl.pca.x %>% 
  ggplot(aes(x = PC1, y = PC2, shape = Folate, color = Status)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC1 (36.0% Var)") +
  ylab("PC2 (16.9% Var)") +
  ggtitle("Principal Component Analysis\nExperimental samples only\nLrp6KO Exp 2 / Positive Mode") +
  scale_color_tableau()
ko2.pos.smpl.pca.x %>% 
  ggplot(aes(x = PC3, y = PC4, shape = Folate, color = Status)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab("PC3 (12.8% Var)") +
  ylab("PC4 (8.6% Var)") +
  ggtitle("Principal Component Analysis\nExperimental samples only\nLrp6KO Exp 2 / Positive Mode") +
  scale_color_tableau()
```


# Batch Effects and Signifiance Testing

## Negative Mode

```{r neg_stat, comment=NA}
# sample prep
ko2.neg.smpl.data <- ko2.neg.noNA %>% 
  filter(Type == "sample") %>% 
  unite("Condition", Folate:Status, sep = "_")
ko2.neg.data.for.sva <- as.matrix(
  ko2.neg.smpl.data[, which(
    colnames(ko2.neg.smpl.data) == "Lrp6KO2n1"
    ):ncol(ko2.neg.smpl.data)]
  )
row.names(ko2.neg.data.for.sva) <- ko2.neg.smpl.data$Samples
ko2.neg.data.for.sva <- t(ko2.neg.data.for.sva)
# pheno prep
ko2.neg.data.pheno <- as.data.frame(ko2.neg.smpl.data[, 4:5])
row.names(ko2.neg.data.pheno) <- ko2.neg.smpl.data$Samples
# sva calculation
ko2.neg.mod <- model.matrix(~ 0 + as.factor(Condition), data = ko2.neg.data.pheno)
ko2.neg.mod0 <- model.matrix(~ 1, data = ko2.neg.data.pheno)
set.seed(2018)
num.sv(ko2.neg.data.for.sva, ko2.neg.mod, method = "be")
set.seed(2018)
num.sv(ko2.neg.data.for.sva, ko2.neg.mod, method = "leek")
set.seed(2018)
ko2.neg.sv <- sva(ko2.neg.data.for.sva, ko2.neg.mod, ko2.neg.mod0)
# extract the surrogate variables
ko2.neg.surr.var <- as.data.frame(ko2.neg.sv$sv)
colnames(ko2.neg.surr.var) <- c("S1")

ko2.neg.covar <- ko2.neg.smpl.pca.x %>% 
  select(Samples, PC1:PC5) %>% 
  inner_join(
    cbind(ko2.neg.data.pheno, ko2.neg.surr.var) %>% 
      rownames_to_column("Samples"),
    by = "Samples"
    ) %>% 
  as_tibble()
ko2.neg.covar %>% 
  ggplot(aes(Condition, S1, color = Condition)) +
  geom_boxplot() +
  geom_jitter(size = 2, alpha = 0.8, width = 0.1)
summary(aov(S1 ~ Condition, data = ko2.neg.covar))
ko2.neg.covar %>% 
  select(Samples:PC5, Condition) %>% 
  gather("PC", "value", PC1:PC5) %>% 
  ggplot(aes(Condition, value)) +
  geom_boxplot() +
  geom_point(size = 2, alpha = 0.8, aes(color = Condition)) +
  facet_wrap(~ PC, scales = "free") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
ko2.neg.covar %>% 
  select(PC1:PC5, S1) %>% 
  ggcorr(label = TRUE)
ko2.neg.covar %>% 
  select(PC1:PC5, S1) %>% 
  ggpairs()
ko2.neg.covar %>% 
  ggplot(aes(PC1, S1)) +
  geom_point(size = 2, alpha = 0.8)

colnames(ko2.neg.mod) <- c("fa_ko", "fa_wt", "ko", "wt")
# combine the full model matrix and the surrogate variables into one
ko2.neg.design.sv <- cbind(ko2.neg.mod, ko2.neg.surr.var)
ko2.neg.cont.mat <- makeContrasts(
  ko_lowFA = ko - wt, 
  ko_highFA = fa_ko - fa_wt,
  FA_diff = (fa_ko - fa_wt) - (ko - wt),
  levels = c("fa_ko", "fa_wt", "ko", "wt", "S1")
  )
# fit the model/design matrix
ko2.neg.eb <- ko2.neg.data.for.sva %>% 
  lmFit(ko2.neg.design.sv) %>% 
  contrasts.fit(ko2.neg.cont.mat) %>% 
  eBayes()
ko2.neg.eb.tidy <- tidy(ko2.neg.eb) %>% 
  mutate(
    adj_pval = p.adjust(p.value, method = "bonferroni"),
    FC = 2 ^ estimate,
    change_in_vpa = ifelse(FC < 1, "down", "up")
    ) %>% 
  rename(compound_short = gene)
ko2.neg.eb.tidy %>% 
  filter(adj_pval < 0.05 & (FC > 1.2 | FC < 1/1.2)) %>% 
  inner_join(ko2.neg.compound.info, by = c("compound_short" = "Compound_short"))
# volcano plot
ko2.neg.eb.tidy %>% 
  ggplot(aes(estimate, -log10(adj_pval))) +
  geom_point(size = 2, alpha = 0.5) +
  geom_hline(linetype = "dashed", color = "#009E73", yintercept = -log10(0.05)) +
  geom_vline(linetype = "dashed", color = "#CC79A7", xintercept = log2(1.2)) +
  geom_vline(linetype = "dashed", color = "#CC79A7", xintercept = log2(1/1.2)) +
  xlab("log2(FC)") +
  ylab("-log10(adjusted p-value)") +
  ggtitle("Volcano plot\nLrp6KO Exp 2 / Negative Mode")
```


## Positive Mode

```{r pos_stat, comment=NA}
# sample prep
ko2.pos.smpl.data <- ko2.pos.noNA %>% 
  filter(Type == "sample") %>% 
  unite("Condition", Folate:Status, sep = "_")
ko2.pos.data.for.sva <- as.matrix(
  ko2.pos.smpl.data[, which(
    colnames(ko2.pos.smpl.data) == "Lrp6KO2p10"
    ):ncol(ko2.pos.smpl.data)]
  )
row.names(ko2.pos.data.for.sva) <- ko2.pos.smpl.data$Samples
ko2.pos.data.for.sva <- t(ko2.pos.data.for.sva)
# pheno prep
ko2.pos.data.pheno <- as.data.frame(ko2.pos.smpl.data[, 4:5])
row.names(ko2.pos.data.pheno) <- ko2.pos.smpl.data$Samples
# sva calculation
ko2.pos.mod <- model.matrix(~ 0 + as.factor(Condition), data = ko2.pos.data.pheno)
ko2.pos.mod0 <- model.matrix(~ 1, data = ko2.pos.data.pheno)
set.seed(2018)
num.sv(ko2.pos.data.for.sva, ko2.pos.mod, method = "be")
set.seed(2018)
num.sv(ko2.pos.data.for.sva, ko2.pos.mod, method = "leek")
set.seed(2018)
ko2.pos.sv <- sva(ko2.pos.data.for.sva, ko2.pos.mod, ko2.pos.mod0)
# extract the surrogate variables
ko2.pos.surr.var <- as.data.frame(ko2.pos.sv$sv)
colnames(ko2.pos.surr.var) <- c("S1", "S2", "S3")

ko2.pos.covar <- ko2.pos.smpl.pca.x %>% 
  select(Samples, PC1:PC5) %>% 
  inner_join(
    cbind(ko2.pos.data.pheno, ko2.pos.surr.var) %>% 
      rownames_to_column("Samples"),
    by = "Samples"
    ) %>% 
  as_tibble()
ko2.pos.covar %>% 
  select(Samples, Condition:S3) %>% 
  gather("surr_var", "value", S1:S3) %>% 
  ggplot(aes(Condition, value, color = Condition)) +
  geom_boxplot() +
  geom_jitter(size = 2, alpha = 0.8, width = 0.1) +
  facet_wrap(~ surr_var) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

ko2.pos.covar %>% 
  select(Samples:PC5, Condition) %>% 
  gather("PC", "value", PC1:PC5) %>% 
  ggplot(aes(Condition, value)) +
  geom_boxplot() +
  geom_point(size = 2, alpha = 0.8, aes(color = Condition)) +
  facet_wrap(~ PC, scales = "free") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
ko2.pos.covar %>% 
  select(PC1:PC5, S1:S3) %>% 
  ggcorr(label = TRUE)
ko2.pos.covar %>% 
  select(PC1:PC5, S1:S3) %>% 
  ggpairs()
ko2.pos.covar %>% 
  ggplot(aes(PC1, S1)) +
  geom_point(size = 2, alpha = 0.8)

colnames(ko2.pos.mod) <- c("fa_ko", "fa_wt", "ko", "wt")
# combine the full model matrix and the surrogate variables into one
ko2.pos.design.sv <- cbind(ko2.pos.mod, ko2.pos.surr.var[, 1])
colnames(ko2.pos.design.sv)[5] <- "S1"
ko2.pos.cont.mat <- makeContrasts(
  ko_lowFA = ko - wt, 
  ko_highFA = fa_ko - fa_wt,
  FA_diff = (fa_ko - fa_wt) - (ko - wt),
  levels = c("fa_ko", "fa_wt", "ko", "wt", "S1")
  )
# fit the model/design matrix
ko2.pos.eb <- ko2.pos.data.for.sva %>% 
  lmFit(ko2.pos.design.sv) %>% 
  contrasts.fit(ko2.pos.cont.mat) %>% 
  eBayes()
ko2.pos.eb.tidy <- tidy(ko2.pos.eb) %>% 
  mutate(
    adj_pval = p.adjust(p.value, method = "BH"),
    FC = 2 ^ estimate,
    change_in_vpa = ifelse(FC < 1, "down", "up")
    ) %>% 
  rename(compound_short = gene)
ko2.pos.eb.tidy %>% 
  filter(adj_pval < 0.05 & (FC > 1.2 | FC < 1/1.2)) %>% 
  inner_join(ko2.pos.compound.info, by = c("compound_short" = "Compound_short"))
```

# Conclusion

Nothing significant - move on

# Session Info

```{r session_info, comment=NA}
sessionInfo()
```

